---
title: "Flat worm plots"
author: "Matthew Kay"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Flat Worm Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

This vignette covers some variaties of worm plots, and in particular introduces flatworm plots, a regression diagnostic plot used to assess residual fit---like a QQ plot, but modified to maximize the ability of humans to perceive deviations from model fit.

# Setup

## Libraries and imports

## Required libraries

If you are missing any of the packages below, use `install.packages("packagename")` to install them.
The `import::` syntax requires the `import` package to be installed, and provides a simple way to 
import specific functions from a package without polluting your entire namespace (unlike `library()`)

```{r setup, results="hide", message=FALSE}
library(dplyr)
library(magrittr)   #pipe syntax (%>%, %<>%, etc)
library(ggplot2)
library(gamlss)
import::from(gamlss.dist, dTF, qTF, pTF, rTF)   #the TF functions are a scaled and shifted t distribution
```

# Test of kurtosis

## Data

Let's generate some data with varying levels of kurtosis

```{r}
k = 100
d = data_frame(
    mu = 0,
    sigma = 1,
    nu = seq(1,50,by=1)
) %>%
    group_by(mu, sigma, nu) %>%
    do(data_frame(
        x = rTF(k, .$mu, .$sigma, .$nu)
    ))
```

And a normal model:

```{r}
dn = data_frame(
    x = rnorm(k, 0, 1)
)
mn = lm(x ~ 1, data = dn)
```

Now let's fit the models:

```{r}
m = d %>%
    group_by(nu) %>%
    do(model = lm(x ~ 1, data=.))
```


```{r}
flatworm(m$model[[10]], ylim=6)
+ geom_vline(xintercept = c(-2,2))
flatworm(mn, cubic=T)
wp(resid = rstandard(m$model[[10]]))
wp(resid = rstandard(mn))
```

# Test of skew

Let's generate some data with varying levels of skew

```{r}
k = 1000
d = data_frame(
    mu = 0,
    sigma = 1,
    skew = seq(-100,100,by=10)
) %>%
    group_by(mu, sigma, skew) %>%
    do(data_frame(
        x = rST1(k, .$mu, .$sigma, .$skew, Inf)
    ))
```

And a normal model:

```{r}
dn = data_frame(
    x = rnorm(k, 0, 1)
)
mn = lm(x ~ 1, data = dn)
```

Now let's fit the models:

```{r}
m = d %>%
    group_by(skew) %>%
    do(model = lm(x ~ 1, data=.))
```

```{r}
flatworm(m$model[[0 + 11]], ylim=6, lines=FALSE, loess=TRUE, points=FALSE) + geom_vline(xintercept = c(-2,2))
wp(resid = rstandard(m$model[[-5 + 11]]))
```

