---
title: "Slab + interval stats and geoms"
author: "Matthew Kay"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Slab + interval stats and geoms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
<style type="text/css">
.kable-table table {
  margin-left: 0;
}
img {
  border: none;
}
</style>

```{r chunk_options, include=FALSE}
tiny_width = 5.5
tiny_height = 3 + 2/3
small_width = med_width = 6.75
small_height = med_height = 4.5
large_width = 8
large_height = 5.25

knitr::opts_chunk$set(
  fig.width = small_width,
  fig.height = small_height
)
if (capabilities("cairo") && Sys.info()[['sysname']] != "Darwin") {
  knitr::opts_chunk$set(
    dev = "png",
    dev.args = list(type = "cairo")
  )
}
```

## Introduction
  
This vignette describes the slab+interval geoms and stats in `ggdist`. This is a flexible family of stats and geoms designed to make plotting distributions (such as priors and posteriors in Bayesian models, or even sampling distributions from other models) straightforward, and support a range of useful plots, including intervals, eye plots (densities + intervals), CCDF bar plots (complementary cumulative distribution functions + intervals), gradient plots, and histograms.

## Setup
  
The following libraries are required to run this vignette:
  
```{r setup, message = FALSE, warning = FALSE}
library(dplyr)
library(tidyr)
library(distributional)
library(ggdist)
library(ggplot2)
library(cowplot)
library(patchwork)

theme_set(theme_ggdist())
```

```{r hidden_options, include=FALSE}
.old_options = options(width = 120)
```


## Roadmap: It all starts with "slabinterval"

```{r slabinterval_family, echo=FALSE, fig.height=5.5, fig.width=6.2}
dists_df = tibble(
  # enforce order
  geom = rev(c(
    "halfeye", 
    "eye",
    "gradientinterval", 
    "ccdfinterval", 
    "cdfinterval",
    "interval",
    "pointinterval",
    "slab",
    "histinterval",
    "dots",
    "dotsinterval"
    )) %>%
    factor(., levels = .),
  dist = dist_normal(4, 1)
)

hist_df = tibble(
  geom = "histinterval",
  x = qnorm(ppoints(1000), 4, 1)
)

dists_df_ = function(geom_) filter(dists_df, geom == geom_)


# FAMILY HEADER

dists_xlim = c(0,8)

header_theme = theme(
  axis.line.x = element_blank(),
  plot.background = element_rect(fill = "gray95"),
  panel.background = element_blank(),
  plot.margin = unit(c(5.5, 0, 5.5, 5.5), "points")
)

dists_header_plot = dists_df_("halfeye") %>%
  mutate(geom = "slabinterval") %>%
  ggplot(aes(y = geom, xdist = dist)) +
  stat_slabinterval(position = position_nudge(y = - 0.2)) +
  scale_x_continuous(limits = dists_xlim, expand = c(0,0), breaks = NULL) +
  scale_y_discrete(expand = c(0.4,0)) +
  labs(
    subtitle = "The stat_slabinterval / geom_slabinterval family",
    x = NULL,
    y = NULL
  ) +
  header_theme

statgeom_theme = list(
  labs(y = NULL, x = NULL),
  theme(
    axis.line.x = element_blank(),
    axis.line.y = element_blank(),
    axis.ticks = element_blank(),
    plot.margin = unit(c(5.5, 5.5, 5.5, 0), "points")
  )
)

statgeom_header_plot = data.frame(
    geom = factor("slabinterval"),
    prefix = factor(c("stat_...", "geom_..."), levels = c("stat_...", "geom_..."))
  ) %>%
  ggplot(aes(x = prefix, y = geom)) +
  geom_hline(aes(yintercept = as.numeric(geom) - .1), color = "gray80", data = . %>% filter(prefix == "stat_...")) +
  geom_point(size = 5, color = "gray65", position = position_nudge(y = -.1)) +
  scale_x_discrete(position = "top") +
  scale_y_discrete(breaks = NULL, expand = c(0.4,0)) +
  statgeom_theme +
  header_theme


# SHORTCUT STATS

dists_plot = dists_df %>%
  ggplot(aes(y = geom, xdist = dist)) +
  geom_blank() + # ensures order
  stat_eye(data = dists_df_("eye")) +
  stat_halfeye(data = dists_df_("halfeye"), position = position_nudge(y = -0.2)) +
  stat_gradientinterval(data = dists_df_("gradientinterval"), scale = .5, fill_type = "gradient") +
  stat_ccdfinterval(data = dists_df_("ccdfinterval"), scale = .5) +
  stat_cdfinterval(data = dists_df_("cdfinterval"), scale = .5) +
  stat_interval(data = dists_df_("interval"), color = "gray65", alpha = 1/3, linewidth = 10,
    position = position_nudge(y = -.1)) +
  stat_pointinterval(data = dists_df_("pointinterval")) +
  stat_slab(data = dists_df_("slab"), position = position_nudge(y = - 0.2)) +
  stat_histinterval(aes(x = x, xdist = NULL), data = hist_df, position = position_nudge(y = - 0.25)) +
  stat_dotsinterval(data = dists_df_("dotsinterval"), position = position_nudge(y = - 0.35)) +
  stat_dots(data = dists_df_("dots"), position = position_nudge(y = - 0.35)) +
  scale_slab_alpha_continuous(guide = "none") +
  scale_x_continuous(limits = dists_xlim, expand = c(0,0)) +
  labs(
    x = NULL,
    y = NULL
  )

statgeom_plot = tribble(
    ~geom,                 ~prefix,
    "halfeye",          c("stat_..."), 
    "eye",              c("stat_..."), 
    "gradientinterval", c("stat_..."), 
    "ccdfinterval",     c("stat_..."), 
    "cdfinterval",      c("stat_..."), 
    "interval",         c("stat_...", "geom_..."),
    "pointinterval",    c("stat_...", "geom_..."),
    "slab",             c("stat_...", "geom_..."),
    "histinterval",     c("stat_..."),
    "dots",             c("stat_...", "geom_..."),
    "dotsinterval",     c("stat_...", "geom_..."),
  ) %>%
  unnest(prefix) %>%
  mutate(
    geom = factor(geom, levels = levels(dists_df$geom)),
    prefix = factor(prefix, levels = c("stat_...", "geom_..."))
  ) %>%
  ggplot(aes(x = prefix, y = geom)) +
  geom_hline(aes(yintercept = as.numeric(geom) - .1), color = "gray80", data = . %>% filter(prefix == "stat_...")) +
  geom_point(size = 5, color = "gray65", position = position_nudge(y = -.1)) +
  scale_x_discrete(breaks = NULL) +
  scale_y_discrete(breaks = NULL, expand = c(0,.6)) +
  statgeom_theme

dists_header_plot + statgeom_header_plot +
  dists_plot + statgeom_plot +
  plot_layout(ncol = 2, widths = c(0.75, 0.25), heights = c(1, 10))
```

`ggdist` has a pantheon of geoms and stats that stem from a common root: `geom_slabinterval()` and `stat_slabinterval()`. These geoms consist of a "slab" (say, a density or a CDF), one or more intervals, and a point summary. These components may be computed in a number of different ways, and different variants of the geom will or will not include all components.

The base `geom_slabinterval()` uses a variety of custom aesthetics to create the composite geometry:

```{r slabinterval_components, echo=FALSE, fig.height=4.2, fig.width=6.5}
red_ = "#d95f02"
green_ = "#1b9e77"
blue_ = "#7570b3"

bracket_ = function(..., x, xend = x, y, yend = y, color = red_) {
  annotate("segment",
    arrow = arrow(angle = 90, ends = "both", length = unit(3, "points")),
    color = color, linewidth = 0.75,
    x = x, xend = xend, y = y, yend = yend,
    ...
  )
}
thickness_ = function(x) dnorm(x,4,1) * 0.9 / dnorm(4,4,1)
thickness_bracket_ = function(x) bracket_(x = x, y = 0, yend = thickness_(x))
refline_ = function(..., x, xend = x, y, yend = y, color = red_, linetype = "solid", alpha = 0.5) {
  annotate("segment",
    color = color, linetype = linetype, alpha = alpha, linewidth = 0.75,
    x = x, xend = xend, y = y, yend = yend,
    ...
  )
}
label_ = function(..., hjust = 0, color = red_) {
  annotate("text",
    color = color, hjust = hjust, lineheight = 1,
    size = 3.25,
    ...
  )
}
arrow_ = function(..., curvature = 0, x, xend = x, y, yend = y) {
  annotate("curve", 
    color = red_, arrow = arrow(angle = 45, length = unit(3, "points"), type = "closed"),
    curvature = curvature,
    x = x, xend = xend, y = y, yend = yend
  )
}


dists_df_("halfeye") %>%
  ggplot(aes(y = 0, xdist = dist)) +
  stat_slabinterval(
    aes(linewidth = NULL),
    slab_color = "black", 
    expand = FALSE, 
    limits = c(0, 8), 
    .width = 1 - 2*pnorm(-1),
    fill = "gray75",
    point_size = 3,
    shape = 21,
    stroke = 1.5,
    linewidth = 3,
  ) +
  
  # height
  refline_(x = 0, xend = 8.4, y = 1) +
  bracket_(x = 8.4, y = 0, yend = 1) +
  label_(label = "height", x = 8.6, y = 1) +
  
  # scale
  refline_(x = 4, xend = 8.6, y = 0.9) +
  bracket_(x = 8.6, y = 0, yend = 0.9) +
  label_(label = "scale = 0.9", x = 8.8, y = 0.9) +
  
  # thickness
  thickness_bracket_(2) +
  thickness_bracket_(2.2) +
  thickness_bracket_(2.4) +
  label_(label = "thickness", hjust = 1, x = 1.63, y = thickness_(2.2), vjust = 0) +
  arrow_(curvature = 0.2, x = 1.67, xend = 1.87, y = thickness_(2.2), yend = thickness_(2) + 0.01) +
  arrow_(x = 1.67, xend = 2.07, y = thickness_(2.2) + 0.01, yend = thickness_(2.2)) +
  arrow_(curvature = -0.2, x = 1.67, xend = 2.27, y = thickness_(2.2) + 0.02, yend = thickness_(2.4)) +
  
  # slab line properties
  label_(x = 2.5, y = 0.7, 
    label = 'slab_color = "black"\nslab_linewidth = 1\nslab_linetype = linetype = "solid"',
    vjust = 1, hjust = 1
  ) +
  arrow_(x = 2.52, xend = 3.08, y = 0.67, yend = thickness_(3.08) + 0.03, curvature = -0.2) +
  
  # slab fill
  label_(x = 5.5, y = 0.7,
    label = 'slab_fill = fill = "gray75"\nslab_alpha = alpha = 1',
    vjust = 1,
  ) +
  arrow_(x = 5.48, xend = 4.5, y = 0.67, yend = thickness_(3), curvature = 0.2) +

  # xmin / x / xmax
  arrow_(x = 2.65, xend = 3, y = -0.1, yend = -0.01, curvature = -0.2) +
  label_(x = 2.7, y = -0.1, label = "xmin", hjust = 1, vjust = 1) +
  arrow_(x = 4, y = -0.1, yend = -0.04) +
  label_(x = 4, y = -0.1, label = "x", hjust = 0.5, vjust = 1) +
  arrow_(x = 5.35, xend = 5, y = -0.1, yend = -0.01, curvature = 0.2) +
  label_(x = 5.3, y = -0.1, label = "xmax", hjust = 0, vjust = 1) +
  
  # interval properties
  label_(x = 3.5, y = -0.2,
    label = 'interval_color = color = "black"\ninterval_alpha = alpha = 1\ninterval_linetype = linetype = "solid"\nlinewidth = size = 3',
    vjust = 1, hjust = 1
  ) +
  arrow_(x = 3.3, xend = 3.4, y = -0.18, yend = -0.01, curvature = -0.1) +

  # point properties
  label_(x = 4.5, y = -0.2,
    label = 'point_fill = fill = "gray75"\npoint_color = color = "black"\npoint_alpha = alpha = 1\npoint_size = size = 3\nshape = 21\nstroke = 1.5',
    vjust = 1, hjust = 0
  ) +
  arrow_(x = 4.55, xend = 4.12, y = -0.18, yend = -0.02, curvature = 0.2) +
  
  coord_cartesian(xlim = c(-1, 10), ylim = c(-0.6, 1)) +
  labs(subtitle = "Properties of geom_slabinterval")
```

Depending on whether you want a horizontal or vertical orientation, you can provide `ymin` and `ymax` instead of `xmin` and `xmax`. By default, some aesthetics (e.g., `fill`, `color`, `size`, `alpha`) set properties of multiple sub-geometries at once. For example, the `color` aesthetic by default sets both the color of the point and the interval, but can also be overridden by `point_color` or `interval_color` to set the color of each sub-geometry separately. 

`geom_slabinterval()` is most useful when paired with `stat_slabinterval()`, which will automatically calculate intervals, densities, and cumulative distribution functions, and maps these onto endpoints of the interval sub-geometry or the `thickness` of the slab sub-geometry.

Using `geom_slabinterval()` and `stat_slabinterval()` directly is not always advisable: they are highly configurable on their own, but this configurability requires remembering a number of combinations of options to use. For quick plotting, ggdist contains a number of pre-configured, easier-to-remember **shortcut stats and geoms** built on top of the slabinterval:

* **Shortcut geoms**, starting with `geom_`, are meant to be used on already-summarized data: `geom_pointinterval()` and `geom_interval()` (for data summarized into intervals) and `geom_slab()` (for data summarized into function values, like densities or cumulative distribution functions).

* **Shortcut stats**, starting with `stat_`, which compute relevant summaries (densities, CDFs, points, and/or intervals) before forwarding the summaries to their geom. Some have geom counterparts (e.g. `stat_interval()` corresponds to `geom_interval()`, except the former applies to sample data and the latter to already-summarized data). Many of these stats do not currently have geom counterparts (e.g. `stat_ccdfinterval()`), as they are primarily differentiated based on what kind of statistical summary they compute. If you've already computed a function (such as a density or CDF), you can just use `geom_slabinterval()` directly. These stats can be used on two types of data, depending on what aesthetic mappings you provide:

   * **Sample data**; e.g. draws from a data distribution, bootstrap distribution, Bayesian posterior distribution (or any other distribution, really). To use the stats on sample data, map sample values onto the `x` or `y` aesthetic. 
   
   * **Distribution objects and analytical distributions**. To use the stats on this type of data, you must use the `xdist`, or `ydist` aesthetics, which take [distributional](https://pkg.mitchelloharawild.com/distributional/) objects, `posterior::rvar()` objects, or distribution names (e.g. `"norm"`, which refers to the Normal distribution provided by the `dnorm/pnorm/qnorm` functions).

All slabinterval geoms can be plotted horizontally or vertically. Depending on how aesthetics are mapped, they will attempt to automatically determine the orientation; if this does not produce the correct result, the orientation can be overridden by setting `orientation = "horizontal"` or `orientation = "vertical"`.

We'll start with one of the most common existing use cases for these kinds geoms: eye plots.

## Eye plots and half-eye plots 

### On sample data

Eye plots combine densities (as violins) with intervals to give a more detailed picture of uncertainty than is available just by looking at intervals. 

For these first few demos we'll use these data:

```{r sample_data}
set.seed(1234)
df = tribble(
    ~group, ~subgroup, ~value,
    "a",          "h", rnorm(1000, mean = 5),
    "b",          "h", rnorm(1000, mean = 7, sd = 1.5),
    "c",          "h", rnorm(1000, mean = 8),
    "c",          "i", rnorm(1000, mean = 9),
    "c",          "j", rnorm(1000, mean = 7)
  ) %>%
  unnest(value)
```

We can summarize it at the group level using an eye plot with `stat_eye()` (ignoring subgroups for now):

```{r group_eye, fig.width = small_height, fig.height = small_height}
df %>%
  ggplot(aes(y = group, x = value)) +
  stat_eye() +
  ggtitle("stat_eye()")
```

Users of older versions of `tidybayes` (which used to contain the `ggdist` geoms) might have used `geom_eye()`, which is the older spelling of `stat_eye()`. Due to the name standardization in version 2 of `tidybayes` (see the description above), `stat_eye()` is now the preferred spelling. `geom_eye()` will continue to work for now, but is deprecated and may throw a warning in future versions.

We can also use `stat_halfeye()` instead to get densities instead of violins:

```{r group_halfeye, fig.width = small_height, fig.height = small_height}
df %>%
  ggplot(aes(y = group, x = value)) +
  stat_halfeye() +
  ggtitle("stat_halfeye()")
```

Or use the `side` parameter to more finely control where the slab (in this case, the density) is drawn:

```{r eye_side, fig.width = med_width, fig.height = small_height}
p = df %>%
  ggplot(aes(x = group, y = value)) +
  panel_border()

plot_grid(ncol = 3, align = "hv",
  p + stat_eye(side = "left") + labs(title = "stat_eye()", subtitle = "side = 'left'"),
  p + stat_eye(side = "both") + labs(subtitle = "side = 'both'"),
  p + stat_eye(side = "right")  + labs(subtitle = "side = 'right'")
)
```

Note how the above chart was drawn vertically instead of horizontally: all slabinterval geoms automatically detect their orientation based on the input data. For example, because we used a factor on the `x` axis above, the geoms were be drawn along the other axis (the `y` axis).
If automatic detection of the desired axis fails, you can specify it manually; e.g. with `stat_halfeye(orientation = 'vertical')` or `stat_halfeye(orientation = 'horizontal')`.

The `side` parameter works for horizontal geoms as well. `"top"` and `"right"` are considered synonyms, as are `"bottom"` and `"left"`; either form works with both horizontal and vertical versions of the geoms:

```{r eyeh_side, fig.width = med_width, fig.height = small_height}
p = df %>%
  ggplot(aes(x = value, y = group)) +
  panel_border()

plot_grid(ncol = 3, align = "hv", 
  # side = "left" would give the same result
  p + stat_eye(side = "left") + ggtitle("stat_eye()") + labs(subtitle = "side = 'bottom'"),
  p + stat_eye(side = "both") + labs(subtitle = "side = 'both'"),
  # side = "right" would give the same result
  p + stat_eye(side = "right") + labs(subtitle = "side = 'top'")
)
```

Eye plots are also designed to support dodging through the standard mechanism of `position = "dodge"`. 
Unlike with geom_violin(), densities in groups that are not dodged (here, 'a' and 'b') have the same area and max width as those in groups that are dodged ('c'):

```{r eye_dodge}
df %>%
  ggplot(aes(x = group, y = value, fill = subgroup)) +
  stat_eye(position = "dodge") +
  ggtitle("stat_eye(position = 'dodge')")
```

Dodging works whether geoms are horizontal or vertical.

### On analytical distributions

The same set of (half-)eye plot stats designed for sample data described above can be used on analytical distributions or distribution vectors by using the `xdist`/`ydist` aesthetics instead of `x`/`y`. These stats accept specifications for distributions in one of two ways:

**Using distribution vectors from the [distributional](https://pkg.mitchelloharawild.com/distributional/) package or `posterior::rvar()` objects**: this format uses aesthetics as follows:

* `xdist`, `ydist`, or `dist`: a distribution vector or `posterior::rvar()` produced by functions such as `distributional::dist_normal()`, `distributional::dist_beta()`, `posterior::rvar_rng()`, etc.

**Using distribution names as character vectors**: this is an older, **soft-deprecated** format included for backwards-compatibility, but generally not recommended in new code. This format uses aesthetics as follows:

* `xdist`, `ydist`, or `dist`: the name of the distribution, following R's naming scheme. This is a string which should have `"p"`, `"q"`, and `"d"` functions defined for it: e.g., "norm" is a valid distribution name because the `pnorm()`, `qnorm()`, and `dnorm()` functions define the CDF, quantile function, and density function of the Normal distribution.
* `args` or `arg1`, ..., `arg9`: arguments for the distribution. If you use `args`, it should be a list column where each element is a list containing arguments for the distribution functions; alternatively, you can pass the arguments directly using `arg1`, ..., `arg9`.

For example, here are a variety of normal distributions describing the same data from the previous section:

```{r dist_data}
dist_df = tribble(
    ~group, ~subgroup, ~mean, ~sd,
    "a",          "h",     5,   1,
    "b",          "h",     7,   1.5,
    "c",          "h",     8,   1,
    "c",          "i",     9,   1,
    "c",          "j",     7,   1
)
```

We can use the `distributional::dist_normal()` function to construct a vector of normal distributions from these means and standard deviations, and map it to the `ydist` aesthetic, which sets the distributions drawn along the `y` axis:

```{r dist_eye_dodge_distributional}
dist_df %>%
  ggplot(aes(x = group, ydist = dist_normal(mean, sd), fill = subgroup)) +
  stat_eye(position = "dodge") +
  ggtitle("stat_eye(position = 'dodge')", "aes(ydist = dist_normal(mean, sd))")
```

Distributional vectors, combined with the `xdist` and `ydist` aesthetics, make it easy to visualize a variety of distributions. E.g., here are some Beta distributions:

```{r beta_stacked}
data.frame(alpha = seq(5, 100, length.out = 10)) %>%
  ggplot(aes(y = alpha, xdist = dist_beta(alpha, 10))) +
  stat_halfeye() +
  labs(
    title = "stat_halfeye()",
    subtitle = "aes(xdist = dist_beta(alpha, 10), y = alpha)",
    x = "Beta(alpha,10) distribution"
  )
```

If you want to plot all of these on top of each other (instead of stacked), you could turn off plotting of the interval to make the plot easier to read using `stat_halfeye(show_interval = FALSE, ...)`. A shortcut for `stat_halfeye(show_interval = FALSE, ...)` is `stat_slab()`. We'll also turn off the fill color with `fill = NA` to make the stacking easier to see, and use outline `color` to show the value of `alpha`:

```{r beta_overplotted_slabh}
data.frame(alpha = seq(5, 100, length.out = 10)) %>%
  ggplot(aes(xdist = dist_beta(alpha, 10), color = alpha)) +
  stat_slab(fill = NA) +
  coord_cartesian(expand = FALSE) +
  scale_color_viridis_c() +
  labs(
    title = "stat_slab()",
    subtitle = "aes(xdist = dist_beta(alpha, 10), color = alpha)",
    x = "Beta(alpha,10) distribution",
    y = NULL
  )
```

Distributional vectors also make it easy to visualize different distribution types simultaneously. For example, if we wished to compare a Student's t distribution and a Normal distribution, we can combine them into a single vector of distributions and plot them (these two distributions are particularly useful as they are often needed for visualizing frequentist confidence distributions---see `vignette("freq-uncertainty-vis")`---and Bayesian priors):

```{r norm_vs_t, fig.width = small_height, fig.height = small_height}
tibble(
  dist = c(dist_normal(0,1), dist_student_t(3, 0, 1))
) %>%
  ggplot(aes(y = format(dist), xdist = dist)) +
  stat_halfeye() +
  ggtitle("stat_halfeye()", "aes(xdist = dist)")
```

The `format()` function in `aes(y = format(dist))` generates a string containing a human-readable name for the distribution for labeling purposes.

### Visualizing priors

A particularly good use of the `xdist`/`ydist` aesthetics is to visualize priors. For example, with `brms` you can specify priors using the `brms::prior()` function, which creates data frames with a `"prior"` column indicating the name of the prior distribution as a string. E.g., one might set some priors on the betas and the standard deviation in a model with something like this:

```{r priors_fake, eval=FALSE}
# NB these priors are made up!
priors = c(
  prior(normal(1, 0.5), class = b),
  prior(gamma(2, 2), class = phi),
  # lb = 0 sets a lower bound of 0, i.e. a half-Normal distribution
  prior(normal(0, 1), class = sigma, lb = 0)
)
priors
```

```{r priors, echo=FALSE}
# we want to avoid a brms dependency, so we fake it above and
# just show the output of brms::prior() here
priors = data.frame(
  prior = c("normal(1, 0.5)", "gamma(2, 2)", "normal(0, 1)"),
  class = c("b", "phi", "sigma"), coef = c("", "", ""),
  group = c("", "", ""),
  resp = c("", "", ""),
  dpar = c("", "", ""),
  nlpar = c("", "", ""),
  lb = c(NA, NA, "0"),
  ub = c(NA, NA, NA)
)
priors
```

The `parse_dist()` function can make it easier to visualize these: it takes in string specifications like those produced by `brms` --- `"normal(0,1)"` and `"lognormal(0,1)"` above --- and translates them into `.dist`, `.args`, and `.dist_obj` columns:

```{r parse_dist}
priors %>%
  parse_dist(prior)
```

Notice that it also automatically translates some common distribution names (e.g. "normal") into their equivalent R function names (`"norm"`). It also creates a `.dist_obj` vector
using `distributional::dist_wrap()`. This distribution vector respects truncation bounds set by
the `lb` and `ub` columns output by `brms::prior()`, as on the half-Normal prior for the `sigma`
parameter. The `.dist_obj` vector can be assigned to the `xdist` or `ydist` aesthetic in ggdist:

```{r prior_dist_halfeyeh}
priors %>%
  parse_dist(prior) %>%
  ggplot(aes(y = paste(class, "~", format(.dist_obj)), xdist = .dist_obj)) +
  stat_halfeye() +
  labs(
    title = "stat_halfeye()",
    subtitle = "with brms::prior() and ggdist::parse_dist() to visualize priors",
    x = NULL,
    y = NULL
  )
```

### Sharing thickness scaling across geometries

In some cases, such as visualizing priors and posteriors, it can be helpful to
use multiple geometries (e.g. one for prior and one for posterior). By default,
normalization/scaling of slab thicknesses is controlled by geometries,
not by a scale function. This allows various functionality not
otherwise possible, such as (1) allowing different geometries to have different
thickness scales and (2) allowing the user to control at what level of aggregation
(panels, groups, the entire plot, etc) thickness scaling is done via the `normalize`
parameter to [geom_slabinterval()].

To override this default behavior and make separate geometries use a shared thickness
scale, add `scale_thickness_shared()` to the plot. The difference is illustrated below:

```{r prior_post, fig.width = med_width, fig.height = small_height * 2/3}
prior_post = data.frame(
 prior = dist_normal(0, 1),
 posterior = dist_normal(0.1, 0.5)
)

separate_scale_plot = prior_post %>%
  ggplot() +
  stat_halfeye(aes(xdist = posterior)) +
  stat_slab(aes(xdist = prior), fill = NA, color = "red") +
  labs(
    title = "prior (slab) + posterior (halfeye)",
    subtitle = "default: no shared thickness scale"
  )

shared_scale_plot = prior_post %>%
  ggplot() +
  stat_halfeye(aes(xdist = posterior)) +
  stat_slab(aes(xdist = prior), fill = NA, color = "#e41a1c") +
  scale_thickness_shared() +
  labs(subtitle = "with scale_thickness_shared()")

separate_scale_plot + shared_scale_plot
```

With `scale_thickness_shared()` applied, both densities have the same area under
their curves.

### Scale transformations of densities

The `stat_slabinterval()` family also adjusts densities appropriately when scale transformations are applied. For example, here is a log-Normal distribution plotted on a log scale:

```{r dist_halfeyeh_log_scale, fig.width = small_height, fig.height = small_height/1.75}
data.frame(dist = dist_lognormal(log(10), 2*log(10))) %>%
  ggplot(aes(xdist = dist)) +
  stat_halfeye() +
  scale_x_log10(breaks = 10^seq(-5,7, by = 2))
```

As expected, a log-Normal density plotted on the log scale appears Normal. The Jacobian for the scale transformation is applied to the density so that the correct density is shown on the log scale. Internally, ggdist attempts to do symbolic differentiation on scale transformation functions (and if that fails, uses numerical differentiation) to calculate the Jacobian so that the `stat_slabinterval()` family works generically across the different scale transformations supported by ggplot.

### Summing up eye plots: `stat_[half]eye`

All of the stats in this section follow the naming scheme `stat_[half]eye`, where adding `half` to the name to yields half-eyes (density plots) instead of eyes (violins).

Like the remaining shortcut stats, these stats also follow these conventions:

* Map sample values to `x` or `y` to use the stats on sample data.
* Use the `xdist`, `ydist`, and `args` aesthetics for analytical distributions or distributions contained in vector objects, such as [distributional](https://pkg.mitchelloharawild.com/distributional/) or `posterior::rvar()` objects.

## Histogram + interval plots

In some cases you might prefer histograms to density plots. `stat_histinterval()` provides an alternative to `stat_halfeye()` that uses histograms instead of densities; it is roughly equivalent to `stat_slabinterval(density = "histogram")`:

```{r stat_histinterval_horizontal, fig.width = med_width, fig.height = small_height}
p = df %>%
  ggplot(aes(x = group, y = value)) +
  panel_border()

ph = df %>%
  ggplot(aes(y = group, x = value)) +
  panel_border()

plot_grid(ncol = 2, align = "hv",
  p + stat_histinterval() + labs(title = "stat_histinterval()", subtitle = "horizontal"),
  ph + stat_histinterval() + labs(subtitle = "vertical")
)
```

You can use the `slab_color` aesthetic to show the outline of the bars. By default the outlines are only drawn along the tops of the bars, as typical tasks with histograms involve area estimation, so the outlines between bars are not strictly necessary and may be distracting. However, if you wish to include those outlines, you can set `outline_bars = TRUE`:

```{r stat_histintervalh_outlines, fig.width = med_width, fig.height = small_height}
plot_grid(ncol = 2, align = "hv",
  ph + stat_histinterval(slab_color = "gray45", outline_bars = FALSE) +
    labs(title = "stat_histinterval", subtitle = "outline_bars = FALSE (default)"),
  ph + stat_histinterval(slab_color = "gray45", outline_bars = TRUE) +
    labs(subtitle = "outline_bars = TRUE")
)
```

### Histograms of analytical distributions

While `stat_histinterval()` will not produce histograms of continuous analytical
distributions, the `stat_slabinterval()` family will automatically detect discrete 
distributions supplied on the `xdist` and `ydist` aesthetics and
plot them using stepped histograms instead of densities. As with `stat_histinterval()`,
you can choose whether or not to draw outlines between bars of the histogram
using `outline_bars = TRUE` or `FALSE` (the default is `FALSE`). 

Here is an example of histograms of analytical distributions that also shows a redundant 
encoding of the density by mapping the `pdf` computed variable onto `fill` (in addition to the default mapping onto `thickness`):

```{r dist_slab_discrete, fig.width = med_width, fig.height = small_height}
tibble(
  group = c("a","b","c","d","e"),
  lambda = c(13,7,4,3,2)
) %>%
  ggplot(aes(x = group)) +
  stat_slab(aes(ydist = dist_poisson(lambda), fill = after_stat(pdf))) +
  geom_line(aes(y = lambda, group = NA), linewidth = 1) +
  geom_point(aes(y = lambda), size = 2.5) +
  labs(fill = "Pr(y)") +
  ggtitle("stat_slab()", "aes(ydist = dist_poisson(lambda), fill = after_stat(pdf))")
```

This was inspired by an example from Isabella Ghement.

## CCDF bar plots

Another (perhaps sorely underused) technique for visualizing distributions is cumulative distribution functions (CDFs) and complementary CDFs (CCDFs). These [can be more effective for some decision-making tasks](https://www.mjskay.com/papers/chi2018-uncertain-bus-decisions.pdf) than densities or intervals, and require fewer assumptions to create from sample data than density plots.

For all of the examples above, both on sample data and analytical distributions, you can replace `[half]eye` with `[c]cdfinterval` to get a stat that creates a CDF or CCDF bar plot. 

`stat_ccdfinterval()` is roughly equivalent to `stat_slabinterval(aes(thickness = after_stat(1 - cdf)), justification = 0.5, side = "topleft", normalize = "none", expand = TRUE)`

### On sample data

`stat_[c]cdfinterval` has the following basic combinations:

```{r cdfinterval_family, fig.width = med_width, fig.height = med_width}
p = df %>%
  ggplot(aes(x = group, y = value)) +
  panel_border()

ph = df %>%
  ggplot(aes(y = group, x = value)) +
  panel_border()

plot_grid(ncol = 2, align = "hv",
  p + stat_ccdfinterval() + labs(title = "stat_ccdfinterval()", subtitle = "vertical"),
  ph + stat_ccdfinterval() + labs(subtitle = "horizontal"),
  p + stat_cdfinterval() + labs(title = "stat_cdfinterval()", subtitle = "vertical"),
  ph + stat_cdfinterval()  + labs(subtitle = "horizontal")
)
```

The CCDF interval plots are probably more useful than the CDF interval plots in most cases, as the bars typically grow up from the baseline. For example, replacing `stat_eye()` with `stat_ccdfinterval()` in our previous subgroup plot produces CCDF bar plots:

```{r ccdf_barplot}
df %>%
  ggplot(aes(x = group, y = value, fill = subgroup, group = subgroup)) +
  stat_ccdfinterval(position = "dodge") +
  ggtitle("stat_ccdfinterval(position = 'dodge')") 
```

The extents of the bars are determined automatically by range of the data in the samples. However, for bar charts it is often good practice to draw the bars from a meaningful reference point (this point is often 0). You can use `ggplot2::expand_limits()` to ensure the bar is drawn down to 0. Let's also adjust the position of the slab relative to the position of the interval using the `justification` parameter:

```{r ccdf_justification}
df %>%
  ggplot(aes(x = group, y = value, fill = subgroup)) +
  stat_ccdfinterval(position = "dodge", justification = 1) +
  expand_limits(y = 0) +
  coord_cartesian(expand = FALSE) +
  ggtitle("stat_ccdfinterval(position = 'dodge', justification = 1)")
```

The `side` parameter also works in the same way it does with `stat_eye()`. Here we'll demonstrate it horizontally:

```{r ccdf_side, fig.width = med_width, fig.height = med_height/1.5}
p = df %>%
  ggplot(aes(x = value, y = group)) +
  expand_limits(x = 0) +
  panel_border()

plot_grid(ncol = 3, align = "hv", 
  # side = "left" would give the same result
  p + stat_ccdfinterval(side = "bottom") + labs(subtitle = "side = 'bottom'") +
    ggtitle("stat_ccdfinterval()"),
  p + stat_ccdfinterval(side = "both") + labs(subtitle = "side = 'both'"),
  # side = "right" would give the same result
  p + stat_ccdfinterval(side = "top") + labs(subtitle = "side = 'top'")
)
```

### On analytical distributions

You can also use `stat_ccdfinterval()` to visualize analytical distributions or distribution vectors, just as you can with `stat_eye()` and `stat_halfeye()`.

By default, `stat_slabinterval()` uses the quantiles at `p = 0.001` and `p = 0.999` of the distributions to determine their extent (unless the lower or upper limit of the distribution's support is finite, in which case that value is used). You can change this setting using the `p_limits` parameter, or use `expand_limits()` to ensure a particular value is shown, as before:

```{r dist_ccdf_dodge}
dist_df %>%
  ggplot(aes(x = group, ydist = dist_normal(mean, sd), fill = subgroup)) +
  stat_ccdfinterval(position = "dodge") +
  expand_limits(y = 0) +
  ggtitle(
    "stat_ccdfinterval(position = 'dodge')",
    "aes(x = dist_normal(mean, sd)) + expand_limits(y = 0)"
  ) +
  coord_cartesian(expand = FALSE)
```

### Summing up CDF bar plots

All of the stats in this section follow the naming scheme `stat_[c]cdfinterval`:

* Add `c` to the name to get CCDFs instead of CDFs.
* Use `xdist`/`ydist` instead of `x`/`y` to use the stats on analytical distributions or distribution vectors instead of sample data.
* It can be helpful to use `expand_limits()` to ensure meaningful reference points are included in the plot.


## Gradient plots

An alternative approach to mapping density onto the `thickness` aesthetic of the slab is to instead map it onto its `alpha` value (i.e., opacity). This is what the `stat_gradientinterval` family does (actually, it uses `slab_alpha`, a variant of the `alpha` aesthetic, described below). 

It is roughly equivalent to `stat_slabinterval(aes(slab_alpha = after_stat(f)), thickness = 1, justification = 0.5)`.

### On sample data

For example, replacing `stat_eye()` with `stat_gradientinterval()` produces gradient + interval plots:

```{r gradient_dodge}
df %>%
  ggplot(aes(x = group, y = value, fill = subgroup)) +
  stat_gradientinterval(position = "dodge") +
  labs(title = "stat_gradientinterval(position = 'dodge')")
```

`stat_gradientinterval()` maps density onto the `slab_alpha` aesthetic, which is a variant of the ggplot `alpha` scale that specifically targets alpha (opacity) values of the slab portion of `geom_slabinterval()`. This aesthetic has default ranges and limits that are a little different from the base ggplot `alpha` scale and which ensure that densities of 0 are mapped onto opacities of 0. You can use `scale_slab_alpha_continuous()` to adjust this scale's settings.

### Avoiding "choppy"-looking gradients

Depending on your graphics device, gradients may be "choppy" looking. 
You can fix this choppiness by setting `fill_type = "gradient"`, which uses a
gradient feature introduced in some graphics engines in R 4.1. If you use
`stat_gradientinterval()` in R 4.1, you will receive a message suggesting you
may want to explicitly set `fill_type = "gradient"` to improve output quality.
If you are using R 4.2 or greater, you should not need to set 
`fill_type = "gradient"` as support for gradients can be auto-detected in that
version, but you will get a warning message if you use `stat_gradientinterval()`
with a graphics engine that does not support gradients.

```{r gradient_dodge_nice}
df %>%
  ggplot(aes(x = group, y = value, fill = subgroup)) +
  stat_gradientinterval(position = "dodge", fill_type = "gradient") +
  labs(title = "stat_gradientinterval(position = 'dodge', fill_type = 'gradient')")
```

As of this writing, in R version 4.1 or greater the graphics devices that support
gradients --- i.e. device that support the `grid::linearGradient()` function ---
include `pdf()`, `svg()`, and `png(type = "cairo")`. See [here](https://developer.r-project.org/Blog/public/2020/07/15/new-features-in-the-r-graphics-engine/) for more about the changes to the R graphics engine.

### On analytical distributions

As with other plot types, you can also use `stat_gradientinterval()` to visualize analytical distributions or distribution vectors:

```{r dist_gradient_dodge}
dist_df %>%
  ggplot(aes(x = group, ydist = dist_normal(mean, sd), fill = subgroup)) +
  stat_gradientinterval(position = "dodge", fill_type = "gradient") +
  labs(
    title = "stat_gradientinterval(position = 'dodge')",
    subtitle = "aes(ydist = dist_normal(mean, sd), fill = subgroup)"
  )
```


## Dotplots

The encodings thus far are *continuous* probability encodings: they map probabilities or probability densities onto aesthetics like `x`/`y` position or `alpha` transparency. An alternative is *discrete* or *frequency-framing* uncertainty visualizations, such as *dotplots* and *quantile dotplots*. *Dotplots* represent distributions by showing each data point, and *quantile dotplots* extend this idea to analytical distributions by showing quantiles from the distribution as a number of discrete possible outcomes.

### On sample data

For example, replacing `stat_halfeye()` with `stat_dots()` produces dotplots. With so many dots here, the outlines mask the fill, so it makes sense to map `subgroup` to the outline `color` of the dots as well:

```{r dots_dodge_nocolor, fig.width = med_width, fig.height = small_height}
df %>%
  ggplot(aes(x = group, y = value, fill = subgroup, color = subgroup)) +
  stat_dots(position = "dodgejust") +
  labs(
    title = "stat_dots()",
    subtitle = "aes(fill = subgroup, color = subgroup))"
  )
```

Unlike the base `ggplot2::geom_dotplot()` geom, `ggdist::geom_dots()` automatically determines a bin width to ensure that the dot stacks fit within the available space. You can set the `binwidth` parameter manually to override this.


### Quantile dotplots

The above plots are a bit hard to read due to the large number of dots. Particularly when summarizing posterior distributions or predictive distributions, which may have thousands of data points, it can make sense to plot a smaller number of dots (say 20, 50 or 100) that are *representative* of the full sample. One such approach is to plot *quantiles*, thereby creating *quantile dotplots*, which can help people make better decisions under uncertainty ([Kay 2016](https://doi.org/10.1145/2858036.2858558), [Fernandes 2018](https://doi.org/10.1145/3173574.3173718)).

The `quantiles` argument to `stat_dots` constructs a quantile dotplot with the specified number of quantiles. Here is one with 50 quantiles, so each dot represents approximately a 2% (1/50) chance. We'll turn off outline color too (`color = NA`):

```{r quantile_dots_dodge, fig.width = med_width, fig.height = small_height}
df %>%
  ggplot(aes(x = group, y = value, fill = subgroup)) +
  stat_dots(position = "dodgejust", quantiles = 50, color = NA) +
  labs(title = "stat_dots(quantiles = 50)")
```

For more on dotplots, see `vignette("dotsinterval")`


## Custom plots

The `slabinterval` family of stats and geoms is designed to be very flexible. Most of the shortcut geoms above can be created simply by setting particular combinations of options and aesthetic mappings using the basic `geom_slabinterval()` and `stat_slabinterval()`. Some useful combinations do not have specific shortcut geoms currently, but can be created manually with only a bit of additional effort.

### Gradients of alpha, color, and fill

Two aesthetics of particular use for creating custom geoms are `slab_alpha`, which changes the alpha transparency of the slab portion of the geom, `slab_color`, which changes its outline color, and `fill`, which changes its fill color. All of these aesthetics can be mapped to variables along the length of the geom (that is, the color does not have to be constant over the entire geom), which allows you to create gradients or to highlight meaningful regions of the data (amongst other things). You can also employ the ggdist-specific `color_ramp` and `fill_ramp`
aesthetics to create custom gradients with outline and fill colors, as demonstrated later in this section.

**Note:** The examples of gradients in this section use the (optional) experimental
setting `fill_type = "gradient"`. If you do not have R greater than 4.1.0 or
are not using a supported graphics device, the output may be blank; in this case,
omit this option. Gradients can be produced without this option but they may
not look as nice.

### CCDF Gradients

By default, `stat_ccdfinterval()` maps the output of the evaluated function (in its case, the CCDF) onto the `thickness` aesthetic of the `slabinterval` geom, which determines how thick the slab is. This is the equivalent of setting `aes(thickness = after_stat(f))`. However, we could instead create a CCDF gradient plot, a sort of mashup of a CCDF barplot and a density gradient plot, by mapping `after_stat(f)` onto the `slab_alpha` aesthetic instead, and setting `thickness` to a constant (1):

```{r ccdf_gradient, fig.width = med_width, fig.height = small_height}
df %>%
  ggplot(aes(x = group, y = value, fill = subgroup)) +
  stat_ccdfinterval(aes(slab_alpha = after_stat(f)), 
    thickness = 1, position = "dodge", fill_type = "gradient"
  ) +
  expand_limits(y = 0) +
  # plus coord_cartesian so there is no space between bars and axis
  coord_cartesian(expand = FALSE) +
  ggtitle("stat_ccdfinterval(thickness = 1)", "aes(slab_alpha = after_stat(f))")
```

If this approach were applied to bins in a histogram, where each bin had some uncertainty
associated with its height, the result would be a so-called *fuzzygram* ([Haber and Wilkinson 1982](https://doi.org/10.1109/MCG.1982.1674217)).

### Highlighting and other combinations

The ability to map arbitrary variables onto fill or outline colors within a slab allows you to easily highlight sub-regions of a plot. 
Taking the earlier example of visualizing priors, we can add a mapping to the `fill` aesthetic to highlight a region of interest, say ±1.5:

```{r norm_vs_t_highlight, fig.width = med_width, fig.height = small_height}
priors = tibble(
  dist = c(dist_normal(0, 1), dist_student_t(3, 0, 1))
) 

priors %>%
  ggplot(aes(y = format(dist), xdist = dist)) +
  stat_halfeye(aes(fill = after_stat(abs(x) < 1.5))) +
  ggtitle("stat_halfeye()", "aes(fill = after_stat(abs(x) < 1.5)))") +
  # we'll use a nicer palette than the default for highlighting:
  scale_fill_manual(values = c("gray85", "skyblue"))
```

We could also combine these aesthetics arbitrarily. Here is a (probably not very useful) eye plot + gradient plot combination, with the portion of the distribution above 1 highlighted:

```{r norm_vs_t_gradient_eye, fig.width = med_width, fig.height = small_height}
priors %>%
  ggplot(aes(y = format(dist), xdist = dist)) +
  stat_eye(aes(slab_alpha = after_stat(f), fill = after_stat(x > 1)), fill_type = "gradient") +
  ggtitle(
    "stat_eye(fill_type = 'gradient')", 
    "aes(slab_alpha = after_stat(f), fill = after_stat(x > 1)))"
  ) +
  # we'll use a nicer palette than the default for highlighting:
  scale_fill_manual(values = c("gray75", "skyblue"))
```

### Mashups with Correll and Gleicher-style gradients

We can also take advantage of the fact that all slabinterval stats also supply `cdf` and `pdf` aesthetics to create charts that make use of both the CDF and the PDF in their aesthetic mappings. For example, we could create [Correll & Gleicher](https://doi.org/10.1109/TVCG.2014.2346298)-style gradient plots by fading the tails outside of the 95% interval in proportion to $|1 - 2F(x)|$ (where $F(x)$ is the CDF):

```{r correll_gradient, fig.width = small_width, fig.height = small_height/2}
priors %>%
  ggplot(aes(y = format(dist), xdist = dist)) +
  stat_gradientinterval(aes(slab_alpha = after_stat(-pmax(abs(1 - 2*cdf), .95))),
    fill_type = "gradient"
  ) +
  scale_slab_alpha_continuous(guide = "none") +
  ggtitle(
    "stat_gradientinterval(fill_type = 'gradient')",
    "aes(slab_alpha = after_stat(-pmax(abs(1 - 2*cdf), .95)))"
  )
```

We could also do a mashup of faded-tail gradients with violin plots by starting with an eye plot and then using the generated `cdf` aesthetic to fade the tails, producing plots like those in [Helske *et al.*](https://arxiv.org/abs/2002.07671):

```{r helske_gradient_eye, fig.width = med_width, fig.height = small_height}
priors %>%
  ggplot(aes(y = format(dist), xdist = dist)) +
  stat_eye(aes(slab_alpha = after_stat(-pmax(abs(1 - 2*cdf), .95))), fill_type = "gradient") +
  scale_slab_alpha_continuous(guide = "none") +
  ggtitle(
    "stat_eye(fill_type = 'gradient')",
    "aes(slab_alpha = after_stat(-pmax(abs(1 - 2*cdf), .95)))"
  )
```

A related idea is one from [Tukey](https://doi.org/10.1080/00031305.1986.10475361): rather than
visually emphasizing where a value is likely, emphasize where it is *unlikely*. While Tukey
used a visual representation showing both pointwise and simultaneous intervals, for this
example we will do something a bit different, inverting the faded-tails function from 
Correll & Gleicher to create bars that "block out" the regions of low likelihood:

```{r tukey_pencils, fig.width = small_height * 1.25, fig.height = small_height}
dist_df %>%
  ggplot(aes(x = group, ydist = dist_normal(mean, sd), fill = subgroup)) +
  stat_slab(
    aes(
      thickness = after_stat(pmax(0, abs(1 - 2*cdf) - .95)), 
      fill_ramp = after_stat(pmax(0, abs(1 - 2*cdf) - .95))
    ),
    side = "both", position = "dodge", fill_type = "gradient"
  ) +
  labs(
    title = 'stat_slab(side = "both")', 
    subtitle = paste0(
      "aes(fill = subgroup,\n       ",
      "fill_ramp and thickness = after_stat(pmax(0, abs(1 - 2*cdf) - .95)))"
    )
  ) +
  guides(fill_ramp = "none") +
  coord_cartesian(expand = FALSE)
```

Thanks to a Jessica Hullman for suggesting the Tukey paper that inspired this idea.

### Densities filled according to intervals

Another common chart type involves filling in the interior of a halfeye plot
according to some intervals. Here, we can use the fact that computed variables
from the interval sub-geometry are made available to the slab sub-geometry and
vice versa. 

For example, within the slab sub-geometry, the `.width` and `level`
computed variables correspond to the smallest intervals that contain the `x`
value at that portion of the slab. Thus, we can map `.width` or `level` onto
the slab fill:

```{r halfeye_filled_intervals, fig.width = med_width, fig.height = small_height}
df %>%
  ggplot(aes(y = group, x = value)) +
  stat_halfeye(aes(fill = after_stat(level))) +
  # na.translate = FALSE drops the unnecessary NA from the legend, which covers
  # slab values outside the intervals. An alternative would be to use 
  # na.value = ... to set the color for values outside the intervals.
  scale_fill_brewer(na.translate = FALSE) +
  labs(
    title = "stat_halfeye()", 
    subtitle = "aes(fill = after_stat(level))",
    fill = "interval"
  )
```

(**Note:** in previous versions of ggdist, using `cut_cdf_qi()` was the recommended
way to achieve this affect. That function still exists for backwards compatibility, 
but mapping `level` or `.width` is now the recommended approach, as it generalizes
to other interval types, such as highest-density intervals --- see later.)

To apply the color scale to all values outside the intervals, one option is to
split `stat_halfeye()` into its constituent parts (`stat_slab()` and 
`stat_pointinterval()`), then include a "100%" interval in `.width`:

```{r halfeye_filled_intervals_2, fig.width = med_width, fig.height = small_height}
df %>%
  ggplot(aes(y = group, x = value)) +
  stat_slab(aes(fill = after_stat(level)), .width = c(.66, .95, 1)) +
  stat_pointinterval() +
  scale_fill_brewer() +
  labs(
    title = "stat_slab()", 
    subtitle = "aes(fill = after_stat(level), .width = c(.66, .95, 1))",
    fill = "interval"
  )
```

If we change the interval type used, the resulting `.width` and `level` computed
variables change correspondingly, allowing us to highlight other types of intervals
besides quantile intervals; for example, highest-density intervals:

```{r halfeye_qi_vs_hdi, fig.width = med_width, fig.height = small_height}
qi_plot = data.frame(dist = dist_beta(10, 2)) %>%
  ggplot(aes(xdist = dist)) + 
  stat_halfeye(aes(fill = after_stat(level)), point_interval = median_qi, .width = c(.5, .8, .95)) + 
  scale_fill_brewer(na.value = "gray95") +
  labs(subtitle = "stat_halfeye(aes(fill = after_stat(level)), point_interval = median_qi)")

hdi_plot = data.frame(dist = dist_beta(10, 2)) %>%
  ggplot(aes(xdist = dist)) + 
  stat_halfeye(aes(fill = after_stat(level)), point_interval = mode_hdci, .width = c(.5, .8, .95)) + 
  scale_fill_brewer(na.value = "gray95") +
  labs(subtitle = "stat_halfeye(aes(fill = after_stat(level)), point_interval = mode_hdci)")

qi_plot /
  hdi_plot
```

### Annotating slabs with spikes

`geom_spike()` and `stat_spike()` make it straightforward to apply custom "spike"
annotations to slabs. The easiest way to use spikes is to use `stat_spike()` and
pass it a numeric position or a function giving numeric position(s) at which
you wish to place a spike (or a list of these). If passed a function, the function
will be applied to the *distributional* or `posterior::rvar()` object used
internally to represent the distribution. 

This means that point estimates (e.g., `mean()`, `median()`, `Mode()`), 
quantiles (`quantile()`), and interval estimates (`qi()`, `hdci()`, `hdi()`) 
can be provided to `stat_spike()` directly. This makes it easy to
modify the previous example to highlight how medians and quantile intervals
differ from modes and highest-density intervals in terms of their densities:

```{r halfeye_qi_vs_hdi_spikes, fig.width = med_width, fig.height = small_height}
qi_plot_spikes = data.frame(dist = dist_beta(10, 2)) %>%
  ggplot(aes(xdist = dist)) + 
  stat_slab(aes(fill = after_stat(level)), point_interval = median_qi, .width = c(.5, .95)) + 
  stat_spike(at = c(median, qi)) + 
  scale_fill_brewer(na.value = "gray95") +
  scale_thickness_shared() +
  labs(subtitle = "stat_slab() + stat_spike(at = c(median, qi))")

hdi_plot_spikes = data.frame(dist = dist_beta(10, 2)) %>%
  ggplot(aes(xdist = dist)) + 
  stat_slab(aes(fill = after_stat(level)), point_interval = mode_hdci, .width = c(.5, .95)) + 
  stat_spike(at = c(Mode, hdci)) +
  scale_fill_brewer(na.value = "gray95") +
  scale_thickness_shared() +
  labs(subtitle = "stat_slab() + stat_spike(at = c(Mode, hdci))")

qi_plot_spikes /
  hdi_plot_spikes
```

Note the use of `scale_thickness_shared()`, which ensures that the `thickness`
values for the slabs and the `thickness` values for the spikes (which determine
their heights) use a shared scale, so they line up correctly.

### Using color ramps for `fill` and `color` aesthetics

`ggdist` supplies `color_ramp` (or `colour_ramp`) and `fill_ramp` aesthetics
which can be used to vary ("ramp") the outline or fill colors smoothly from a 
base color (default `"white"`) to whatever color the geometry would otherwise 
have.

Taking the above example with interval-filled slabs, we could use the `fill_ramp` aesthetic
instead of the `fill` aesthetic to set the slab color based on the interval it is in.
We could then vary the base fill color separately from the interval based on 
another column in the original data table, such as the `subgroup` column:

```{r halfeye_filled_intervals_subgroup, fig.width = med_width, fig.height = small_height}
df %>%
  ggplot(aes(y = group, x = value)) +
  stat_halfeye(
    aes(fill = subgroup, fill_ramp = after_stat(level)),
    .width = c(.50, .80, .95),
    # NOTE: we use position = "dodgejust" (a dodge that respects the 
    # justification of intervals relative to slabs) instead of 
    # position = "dodge" here because it ensures the topmost slab does
    # not extend beyond the plot limits
    position = "dodgejust",
  ) +
  # a range from 1 down to 0.2 ensures the fill goes dark to light inside-out
  # and doesn't get all the way down to white (0) on the lightest color
  scale_fill_ramp_discrete(na.translate = FALSE) +
  labs(
    title = "stat_halfeye(position = 'dodgejust')", 
    subtitle = "aes(fill = subgroup, fill_ramp = after_stat(level))",
    fill_ramp = "interval"
  )
```

We could similarly use `stat_interval()` 
with the `color_ramp` aesthetic to vary subgroup color separately from the
whiteness of the intervals. Here, `level` is a variable generated by all stats in
the `stat_slabinterval()` family which contains the level of the generated intervals,
as an ordered factor.

```{r dist_interval_color_ramp, fig.width = med_width, fig.height = small_height}
dist_df %>%
  ggplot(aes(x = group, ydist = dist_normal(mean, sd), color = subgroup)) +
  stat_interval(aes(color_ramp = after_stat(level)), position = "dodge") +
  labs(
    title = "stat_interval()", 
    subtitle = "aes(color = subgroup, color_ramp = after_stat(level))"
  )
```

See `help("scale_color_ramp")` for more information on the color ramp aesthetics/scales.

### Raindrop plots

[Barrowman and Myers](https://doi.org/10.1198/0003130032369) proposed an alternative to density-based eye plots (such
as created by `stat_eye()` by default) called *raindrop plots*.
In these, the thickness of the slab is proportional to `log(pdf)` instead of `pdf`, and
is bounded within the 95% interval. We can construct a function that uses the `pdf` 
and `.width` computed variables to give a thickness proportional to `log(pdf)` 
within the 95% interval, and use it to create raindrop plots.

Barrowman and Myers apply this technique with a 95% raindrop superimposed on
a 99% raindrop, which we can replicate:

```{r raindrop, fig.width = med_width, fig.height = small_height/1.5}
priors %>%
  ggplot(aes(y = format(dist), xdist = dist)) +
  # must also use normalize = "groups" because min(log(pdf)) will be different for each dist
  stat_slab(
    aes(thickness = after_stat(ifelse(.width <= 0.99, log(pdf), NA))), 
    normalize = "groups", fill = "gray85", .width = .99, side = "both"
  ) +
  stat_eye(
    aes(thickness = after_stat(ifelse(.width <= 0.95, log(pdf), NA))), 
    normalize = "groups"
  ) +
  ggtitle(
    'stat_eye(normalize = "groups")',
    "with aes(thickness = after_stat(ifelse(.width <= 0.95, log(pdf), NA)))\nand aes(thickness = after_stat(ifelse(.width <= 0.99, log(pdf), NA)))"
  )
```

### Creating ridge plots

When plotting densities (as in `stat_halfeye()`, `stat_slab()`, etc) it can be useful to
overplot many densities simultaneously, an approach sometimes called *ridge plots*
(e.g. as in the [ggridges](https://wilkelab.org/ggridges/) package).
This can be done by setting `scale` or `height` to a value greater than 1. 
Setting `height` is often the best approach as it will correctly adjust plot boundaries
(unless you need to use `position = "dodge"`, in which case you should use `scale`
and adjust plot boundaries manually).

```{r slab_ridge, fig.width = med_width, fig.height = small_height}
set.seed(1234)

ridges_df = data.frame(
  group = letters[7:1],
  x = rnorm(700, mean = 1:7, sd = 2)
)

ridges_df %>%
  ggplot(aes(y = group, x = x)) +
  stat_slab(height = 2, color = "black") +
  ggtitle("stat_slab(height = 2, color = 'black')")
```

Depending on if it makes sense for your data (for example, if the scale is unbounded),
you may also wish to adjust the `density` and `trim` parameters. The default `density`, 
`density_bounded()`, estimates the bounds of the distribution, which is useful if your data 
has natural boundaries (e.g., is restricted to be positive). But if you know the underlying 
distribution is unbounded, you can set `density = "unbounded"`. You may also want to set
`trim` to `FALSE` to ensure the densities smoothly go down to 0, rather than being cut off
at the limits of the raw data. Combining both of these with `expand = TRUE`
will make each slab expand itself to the limits of the `x` axis.

We'll use `density`, `trim`, and `expand` along with a combination of `fill` and `fill_ramp`
to give each group on the y axis a different color and to vary the fill along
the `x` axis in a way that provides a "softer" form of region of practical equivalence:

```{r slab_ridge_ramp, fig.width = med_width, fig.height = small_height}
ridges_df %>%
  ggplot(aes(
    y = group, x = x, 
    fill = group, fill_ramp = after_stat(abs(x)), 
    color_ramp = after_stat(-dnorm(x, 0, 2))
  )) +
  stat_slab(
    height = 2, color = "gray15", 
    expand = TRUE, trim = FALSE, density = "unbounded",
    fill_type = "gradient", 
    show.legend = FALSE
  ) +
  geom_vline(xintercept = 0, color = "gray85", linetype = "dashed") +
  ggtitle(
    'stat_slab(height = 2, color = "black", expand = TRUE, trim = FALSE)',
    'aes(fill = group, fill_ramp = after_stat(abs(x)), color_ramp = after_stat(-dnorm(x, 0, 2)))'
  ) +
  scale_fill_viridis_d()
```

We use a tighter ramp on `color` compared to `fill` (via `-dnorm()` instead of `abs()`)
because we want the outlines to quickly ramp back to black outside of 0 so that they 
have sufficient contrast against the slabs when they overlap.

### Varying side, scale, and justification within geoms

The `side`, `scale`, and `justification` parameters can also be varied within
all geoms in the `geom_slabinterval()` family, allowing (for example) different groups to hang above or below the interval:

```{r varying_side_dotplot, fig.width = med_width, fig.height = small_height}
dist_df %>%
  filter(subgroup == "h") %>%
  mutate(side = c("top", "both", "bottom")) %>%
  ggplot(aes(y = group, xdist = dist_normal(mean, sd), side = side)) +
  stat_dotsinterval(scale = 2/3) +
  labs(
    title = 'stat_dotsinterval(scale = 2/3)', 
    subtitle = 'aes(xdist = dist_normal(mean, sd), side = c("top","both","bottom"))'
  ) +
  coord_cartesian()
```

### Multiple slabs and intervals in composite plots

Sometimes you may want to include multiple different types of slabs in the same plot
in order to take advantage of the features each slab type provides. For example,
people often combine densities with dotplots to show the underlying datapoints that
go into a density estimate, creating so-called "rain cloud" plots. 

To use multiple
slab geometries together, you can use the `side` parameter to change which side
of the interval a slab is drawn on and set the `scale` parameter to something around
`0.5` (by default it is `0.9`) so that the two slabs do not overlap.
Geoms can also be dodged together, as in this example using densities with
quantile dotplots in subgroups. This example also shows how `stat_pointinterval()`
can be repurposed to be used with other geoms; here to replace points with labels
(the idea of replacing points with labels comes from Brenton Wiernik).

```{r halfeye_quantile_dotplot, fig.width = med_width, fig.height = small_height}
df %>%
  ggplot(aes(x = group, y = value, fill = subgroup)) +
  stat_slab(side = "left", scale = 0.5, position = "dodge") +
  stat_dotsinterval(scale = 0.5, quantiles = 100, position = "dodge") +
  stat_pointinterval(
    geom = "label", 
    aes(label = paste0(group, subgroup)), 
    .width = .5,  # set to a scalar to draw only one label instead of two
    position = position_dodge(width = 1),
    size = 3.5
  ) +
  labs(title = paste0(
    'stat_halfeye(side = "left") +\n',
    'stat_dotsinterval(quantiles = 100) +\n',
    'stat_pointinterval(geom = "label")'
  ))
```

When constructing composite plots it may be useful to position the slab and interval parts of the geometry separately.
While some relative positioning of these geometries is supported by manipulating the `justification` parameter, if you
want complete, separate control over positioning of intervals versus slabs, the simplest approach can be to specify those
geometries separately. 

For example, the following uses a separate specification of a `stat_slab()` and a `stat_pointinterval()` instead of a combined `stat_slabinterval()` in order to use `position_dodge()` on the intervals but not the slabs:

```{r slab_and_pointinterval, fig.width = med_width, fig.height = small_height/1.5}
df %>%
  ggplot(aes(fill = group, color = group, x = value)) +
  stat_slab(alpha = .3) +
  stat_pointinterval(position = position_dodge(width = .4, preserve = "single")) +
  labs(
    title = "stat_slab() and stat_pointinterval()", 
    subtitle = "with position_dodge() applied to the intervals",
    y = NULL
  ) +
  scale_y_continuous(breaks = NULL)
```

(Thanks to Brenton Wiernik for [this example](https://github.com/mjskay/ggdist/issues/27#issuecomment-725578437).)


```{r reset_options, include=FALSE}
options(.old_options)
```
