---
title: "Thickness scale"
author: "Matthew Kay"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Thickness scale}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, child="children/chunk_options.txt"}
```

## Introduction
  
This vignette describes the `thickness` scale / aesthetic used by the slab+interval 
family of geoms and stats in `ggdist` (see `vignette("slabinterval")` for more on 
that family). 

## Setup
  
The following libraries are required to run this vignette:
  
```{r setup, message = FALSE, warning = FALSE}
library(ggplot2)
library(dplyr)
library(distributional)
library(ggdist)

theme_set(theme_ggdist())
```

```{r hidden_options, include=FALSE}
.old_options = options(width = 110)
```

## The `thickness` scale

The `thickness` scale is a positional subscale used by `geom_slabinterval()`
to construct *slabs*, which are ribbon-like geometries with a fixed baseline and a 
height determined by the `thickness` aesthetic. The `thickness` scale is most 
typically used via `stat_slabinterval()` and its derivative stats, which allow you
to map distribution functions (like densities or CDFs) onto the `thickness` aesthetic.

`thickness` exists as a scale separate from the typical `x` and `y` aesthetics
(or even `width`/`height`, `xmin`/`xmax`, or `ymin`/`ymax`) so that it is easy to plot multiple 
slabs on the same plot, each with its own separate `thickness` subscale.
For example:

```{r intro}
df = tibble(
  h = c(0,1,1,0, 0,2,2,0, 0,4/3,4/3,0,  0,1,1,0,     0,1.25,1.25,0, 0,.5,.5,0),
  x = c(0,1,4,5, 4,5,6,7, 6,7,9,10,     11,12,15,16, 1,2,4.2,5.2,   4,5,12,13),
  y     = rep(c("a", "b", "a", "a", "b", "a"), each = 4),
  group = rep(c("c", "c", "d", "d", "c", "c"), each = 4),
  panel = rep(c("e", "e", "e", "e", "f", "f"), each = 4),
  name  = rep(c(1,    2, "3a","3b",  4,   5),  each = 4)
)

df_group = df %>%
  summarise(x = mean(x), .by = c(name, group, panel, y))

df %>%
  ggplot(aes(x = x, y = y, fill = group, thickness = h)) +
  geom_slab(color = "gray25", alpha = 0.75) +
  scale_y_discrete(expand = expansion(add = 0.1)) +
  scale_fill_brewer(palette = "Set2") +
  facet_grid(cols = vars(panel), labeller = "label_both") +
  labs("geom_slab() with default thickness scaling") +
  theme(
    legend.position = "bottom",
    panel.grid.major.y = element_line(color = "gray85"),
    panel.background = element_rect(color = "gray70", fill = NA)
  )
```

In the above plot, the values assigned to the `thickness` aesthetic are
all scaled onto a common scale, and the slabs are positioned with their
baselines at the `y` value specified for each slab.

## Adjusting normalization within one geometry

The default thickness scaling shown above corresponds to the `normalize = "all"`
option of `geom_slab()`. It uses the same scale for all slabs, scaling them
according the the maximum height of the tallest slab.

The `normalize` parameter provides a set of options that define a hierarchy of
increasingly more specific sets within which to scale the slabs:

- `"all"` scales according to the maximum over **all slabs**;
- `"panels"` scales according to the maximum within each **panel** (facet);
- `"xy"` scales according to the maximum within each **x/y position** within each **panel**; and
- `"groups"` scales according to the maximum within each **group** within each **x/y position** within each **panel**.

The **x/y position** referred to above is the unique *x* **or** *y* position of the slab on its
off-axis. For example, this would be the unique *y* positions in the above chart (`"a"` or `"b"`), 
because the chart is drawn with a horizontal `orientation`.

### `normalize = "all"`

We can see how normalization affects the scales by annotating the previous plot
of the default normalization settings:

```{r normalize_all}
subguide_orangered = subguide_outside(
  title = "thickness",
  position = "right",
  theme = theme_ggdist() + theme(
    text = element_text(color = "orangered"),
    axis.line.y = element_line(color = "orangered"),
    axis.ticks.y = element_line(color = "orangered"),
    axis.text.y = element_text(color = "orangered")
  )
)

plot_slabs_with_scales = function(..., subguide = subguide_orangered) {
  df %>%
    ggplot(aes(x = x, y = y, fill = group)) +
    geom_hline(yintercept = c(1,1.9, 2,2.9), color = "orangered", linetype = "11", linewidth = 0.5) +
    geom_slab(
      aes(thickness = h),
      subguide = subguide,
      alpha = 0.75,
      color = "gray25",
      ...
    ) +
    geom_label(
      aes(label = name),
      data = df_group,
      color = "gray25",
      alpha = 0.75,
      vjust = 0,
      show.legend = FALSE
    ) +
    scale_y_discrete(expand = expansion(add = 0.1)) +
    scale_fill_brewer(palette = "Set2") +
    facet_grid(cols = vars(panel), labeller = "label_both") +
    theme(
      plot.margin = margin(5.5, 50, 5.5, 5.5),
      panel.spacing.x = unit(40, "pt"),
      legend.position = "bottom",
      panel.background = element_rect(color = "gray70", fill = NA)
    )
}

plot_slabs_with_scales(normalize = "all") +
  labs(
    title = "geom_slab(normalize = 'all')",
    subtitle = "default normalization settings"
  )
```

Notice how the minimum and maximum value on each `thickness` subscale is the
same: it is the maximum thickness value over all slabs; i.e. the height
of the tallest slab in the data. This is a conservative default setting that 
ensures all slabs in the same geometry are scaled together.

### `normalize = "panels"`

Sometimes you may have separate panels within which you want to scale all slabs
together. You can do this using `normalize = "panels"`:

```{r normalize_panels}
plot_slabs_with_scales(normalize = "panels") +
  labs(
    title = "geom_slab(normalize = 'panels')",
    subtitle = "scales to the maximum height within each panel"
  )
```

Notice how the `thickness` scales *inside* a given panel are the same: panel `"e"`
has a maximum `thickness` of `2` (the height of slab `2`) and panel `"f"` has
a maximum `thickness` of `1.25` (the height of slab `4`). All other slabs within
the same panel are scaled accordingly.

### `normalize = "xy"`

Often it is useful to scale slabs on the same `x` or `y` position together, where
whether we scale within `x` or `y` is determined by the `orientation` of the
geometry. You can do this using `normalize = "xy"`:

```{r normalize_xy}
plot_slabs_with_scales(normalize = "xy") +
  labs(
    title = "geom_slab(normalize = 'xy')",
    subtitle = "scales to the maximum height within each x/y position within each panel"
  )
```

In this plot, because `orientation = "horizontal"`, `normalize = "xy"` scales
within each unique *y* position; i.e. the values `"a"` and `"b"`. Normalization
settings are hierarchical, so it uses the maximum `thickness` value within
each *y* position within each panel. Thus:

- slab `2` is scaled to its maximum of `2.0`;
- slabs `1`, `3a`, and `3b` are scaled to their maximum of `1.33` (the height of slab `3a`);
- slab `4` is scaled to its maximum of `1.25`; and
- slab `5` is scaled to its maximum of `0.5`.

### `normalize = "groups"`

Finally, it can be useful to scale all slabs according to their maximum height. 
You can do this using `normalize = "groups"`. 

Notably, if you have multiple groups at the same x/y position, you cannot 
combine `normalize = "groups"` with any thickness subguide except `"none"`, 
because the thickness subguides on each y value will not be unique. Hence
this error:

```{r normalize_groups_error, error = TRUE}
plot_slabs_with_scales(normalize = "groups")
```

Thus for this example we will have to omit the `thickness` subguide:

```{r normalize_groups}
plot_slabs_with_scales(normalize = "groups", subguide = "none") +
  labs(
    title = "geom_slab(normalize = 'groups')",
    subtitle = "scales to the maximum height within each group within each x/y position within each panel"
  )
```

Notice how all groups are scaled to have their maximum height at the maximum
thickness value; thus, slab `1` and slab `3b` are no longer the same height.

Why is slab `3b` not scaled to touch the maximum `thickness`? `geom_slab()` 
technically only considers values in different groups to be in different slabs,
so slab `3a` and `3b` are actually parts of the same slab, as indicated by the
unbroken outline they are drawn with.

## Sharing `thickness` scales across multiple geometries

Sometimes we want to use different geometries to plot slabs on the same scale.
This often happens when plotting priors and posteriors, but may happen in other
cases as well.

Here is a prior and posterior plotted using separate geometries, with their
thickness scales annotated:

```{r prior_post}
df_prior_post = data.frame(
  prior = dist_normal(0, 1),
  posterior = dist_normal(0.1, 0.3)
)

prior_post_plot = df_prior_post %>%
  ggplot() +
  stat_slab(
    aes(xdist = posterior),
    subguide = subguide_inside(title = "posterior thickness")
  ) +
  stat_slab(
    aes(xdist = prior),
    color = "orangered",
    fill = NA,
    subguide = subguide_orangered(title = "prior thickness", just = 0, label_side = "inside")
  ) +
  scale_y_continuous(breaks = NULL) +
  theme(panel.background = element_rect(color = "gray70", fill = NA))

prior_post_plot +
  labs(
    title = "two different slab geometries",
    subtitle = "thickness scales are not shared across geometries by default"
  )
```

Notice how the two geometries have different `thickness` scales. If we add `scale_thickness_shared()`
to the plot, they will be given the same scale:

```{r prior_post_shared}
prior_post_plot +
  scale_thickness_shared() +
  labs(
    title = "two different slab geometries",
    subtitle = "using scale_thickness_shared() to share thickness scale across geometries"
  )
```

`scaled_thickness_shared()` works by scaling values and then tagging them with
a special `thickness()` datatype. That type carries information about the 
original scale limits (used to draw subguides), and also tells the slab 
geometries that they should not do any further normalization. Thus, when you 
use `scale_thickness_shared()`, the `normalize` parameter on each slab geometry
is ignored.

## Spacing slabs with `height` and `scale`

The `thickness` aesthetic is drawn within a subset of the bounding box of 
`geom_slab()` determined by the `scale` aesthetic, which defaults to `0.9`:

```{r default_height_scale}
cap = arrow(angle = 90, length = unit(5, "pt"), ends = "both")

scale_plot = function(scale = 0.9) {
  tibble(d = dist_normal(c(4,5)), y = c("a","b")) %>%
    ggplot(aes(xdist = d, y = y)) +
    geom_hline(
      yintercept = c(1, 1 + scale, 2, 2 + scale),
      color = "orangered", linetype = "11", linewidth = 0.5
    ) +
    stat_slab(scale = scale, subguide = subguide_orangered) +
    annotate("segment", x = c(-5,-4), xend = c(-5,-4), y = c(1,2), yend = c(2,3),
      arrow = cap, linewidth = 0.75, color = "gray25"
    ) +
    annotate("label", x = c(-5,-4), y = c(1.5, 2.5),
      label = "height = 1", hjust = -0.05, color = "gray25"
    ) +
    annotate("segment", x = c(-2,-1), xend = c(-2,-1), y = c(1,2), yend = c(1,2) + scale,
      arrow = cap, linewidth = 0.75, color = "gray25"
    ) +
    annotate("label", x = c(-2,-1), y = c(1,2) + scale/2,
      label = paste("scale =", scale), hjust = -0.05, color = "gray25"
    ) +
    scale_y_discrete(expand = expansion(add = 0)) +
    scale_x_continuous(limits = c(-5, 10)) +
    theme(
      plot.margin = margin(5.5, 40, 5.5, 5.5),
      panel.background = element_rect(color = "gray70")
    ) +
    coord_cartesian(clip = "off")
}

scale_plot(scale = 0.9) +
  labs(
    title = "geom_slab()",
    subtitle = "using default height = 1 and scale = 0.9"
  )
```

This allows us to adjust the spacing between slabs using `scale`:

```{r scale_reduced}
scale_plot(scale = 0.65) +
  labs(
    title = "geom_slab()",
    subtitle = "using scale = 0.65 to increase spacing between slabs"
  )
```

### Spacing dodged slabs

`height` and `scale` are particularly useful when combined with dodging, as modifying `height`
allows you to change the spacing **between** sets of slabs with the same y value, 
and modifying `scale` allows you to change the spacing **within** sets:

```{r default_dodge, fig.width = 7, fig.height = 7}
dodged_scale_plot = function(height = 1, scale = 0.9, add = 0) {
  baselines = c(1 - height/2, 1, 2 - height/2, 2)
  tibble(
    d = dist_normal(c(4,5,4,5)),
    y = c("a","a","b","b"),
    group = c("d","e","d","e")
  ) %>%
    ggplot(aes(xdist = d, y = y, fill = group)) +
    geom_hline(yintercept = c(1,2) + height/2, color = "gray85") +
    geom_hline(
      yintercept = c(baselines, baselines + height/2*scale),
      color = "orangered",
      linetype = "11",
      linewidth = 0.5
    ) +
    stat_slab(
      height = height, scale = scale,
      subguide = subguide_orangered, position = "dodgejust",
      alpha = 0.75
    ) +
    annotate("segment",
      x = c(-5,-4), xend = c(-5,-4),
      y = c(1,2) - height/2, yend = c(1,2) + height/2,
      arrow = cap, color = "gray25", linewidth = 0.75
    ) +
    annotate("label", x = c(-5,-4), y = c(1, 2),
      label = paste("height =", height), hjust = -0.05,
      color = "gray25"
    ) +
    annotate("segment", x = c(-2,-2,-1,-1), xend = c(-2,-2,-1,-1),
      y = baselines, yend = baselines + scale * height/2,
      arrow = cap, color = "gray25", linewidth = 0.75
    ) +
    annotate("label", x = c(-2,-2,-1,-1),
      y = baselines + scale*height/4, label = paste("scale =", scale),
      hjust = -0.05, color = "gray25"
    ) +
    scale_y_discrete(expand = expansion(add = add)) +
    scale_x_continuous(limits = c(-5, 10)) +
    scale_fill_brewer(palette = "Set2") +
    theme(
      plot.margin = margin(5.5, 40, 5.5, 5.5),
      panel.background = element_rect(color = "gray70"),
      legend.position = "bottom"
    ) +
    coord_cartesian(clip = "off")
}

dodged_scale_plot(height = 1, scale = 0.9, add = 0) +
  labs(
    title = 'geom_slab(position = "dodgejust")',
    subtitle = "using default height = 1 and scale = 0.9"
  )
```

If we decrease `height`, it increases the space **between** sets of slabs with the same *y* value:

```{r dodge_height_reduced, fig.width = 7, fig.height = 7}
dodged_scale_plot(height = 0.8, scale = 0.9, add = 0.6) +
  labs(
    title = 'geom_slab(position = "dodgejust")',
    subtitle = "using height = 0.8 to increase the space between sets"
  )
```

And if we decrease `scale`, it increases the spacing **within** sets of slabs with the
same *y* value:

```{r dodge_height_scale_reduced, fig.width = 7, fig.height = 7}
dodged_scale_plot(height = 0.8, scale = 0.65, add = 0.6) +
  labs(
    title = 'geom_slab(position = "dodgejust")',
    subtitle = "using height = 0.8 and scale = 0.65"
  )
```

### Ridgeline plots with `height > 1`

Finally, `height` can also be greater than `1`, which can be used to create 
overlapping slabs, as in so-called ridgeline plots:

```{r ridgeline}
data.frame(
  d = dist_normal(10:1/4, 1 + 10:1/15),
  y = letters[1:10]
) %>%
  ggplot(aes(xdist = d, y = y)) +
  stat_slab(height = 3, color = "gray25") +
  labs(
    title = "geom_slab()",
    subtitle = "using height > 1 to create ridgeline plots"
  )
```

```{r reset_options, include=FALSE}
options(.old_options)
```
