---
title: "Using Tidy Data with Bayesian Samplers"
author: "Matthew Kay"
date: "`r Sys.Date()`"
output: 
    rmarkdown::html_vignette:
        toc: TRUE
vignette: >
  %\VignetteIndexEntry{Using Tidy Data with Bayesian Samplers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

This vignette introduces the `tidybayes` package, which facilitates the use of tidy data (one observation
per row) with Bayesian samplers in R. This vignette is geared towards working with tidy data in general-purpose samplers like JAGS or Stan. For a similar introduction to the use of `tidybayes` with high-level modelling functions such as those in `brms` or `rstanarm`, see also <code>[vignette("tidy-rstanarm")](tidy-rstanarm.html)</code>.

The default output (and sometimes input) data formats of popular samplers
like JAGS and Stan often don't quite conform to the ideal of [tidy data](http://dx.doi.org/10.18637/jss.v059.i10). For example, input formats
might expect a list instead of a data frame, and for all variables to be encoded as numeric values
(requiring translation of factors to numeric values and the creation of index variables to store
the number of levels per factor or the number of observations in a data frame). Output formats will
often be in matrix form (requiring conversion for use with libraries like ggplot), and will use
numeric indices (requiring conversion back into factor level names if the you wish to make meaningfully-labelled
plots or estimates). `tidybayes` automates all of these sorts of tasks.

### Philosophy

There are a few core ideas that run through the `tidybayes` API that should (hopefully) make it easy to use:

1. __Tidy data does not always mean all parameter names as values__. In contrast to the `ggmcmc` library (which translates sampler results into a data frame with a `Parameter` and `value` column), the `spread_samples` function in `tidybayes` produces in data frames where the columns are named after parameters and (in some cases) indices of those parameters, as automatically as possible and using a syntax as close to the same way you would refer to those variables in the sampler's language as possible. A similar function to `ggmcmc`'s approach is also provided in `gather_samples`, since sometimes you *do* want parameter names as values in a column. The goal is for `tidybayes` to do the tedious work of figuring out how to make a data frame look the way you need it to, including turning parameters with indices like `"b[1,2]"` and the like into tidy data for you.

2. __Fit into the tidyverse and broom__. `tidybayes` methods fit into a workflow familiar to users of the `tidyverse` (`dplyr`, `tidyr`, `ggplot2`, etc), and the `tidy` function from the `broom` package, which means fitting into the pipe (`%>%`) workflow, using and respecting grouped data frames (thus `spread_samples` and `gather_samples` return results already grouped by parameter indices, and methods like `mean_qi` calculate estimates and intervals for parameters and groups simultaneously), using the same output column names that the `broom::tidy` function does where possible, and not reinventing too much of the wheel if it is already made easy by functions provided by existing `tidyverse` packages (unless it makes for much clearer code for a common idiom).

3. __Focus on composable operations and plotting primitives, not monolithic plots and operations__. Several other packages (notably `bayesplot` and `ggmcmc`) already provide an excellent variety of pre-made methods for plotting Bayesian results. `tidybayes` shies away from duplicating this functionality. Instead, it focuses on providing composable operations for generating and manipulating Bayesian samples in a tidy data format, and graphical primitives for `ggplot` that allow you to build custom plots easily. Most simply, where `bayesplot` and `ggmcmc` tend to have functions with many options that return a full ggplot object, `tidybayes` tends towards providing primitives (like `geom`s) that you can compose and combine into your own custom plots. I believe both approaches have their place.

4. __Sensible defaults make life easy.__ But options (and the data being tidy in the first place) make it easy to go your own way when you need to.

5. __Variable names in models should be descriptive, not cryptic__. This principle implies avoiding cryptic (and short) subscripts in favor of longer (but descriptive) ones. This is a matter of readability and accessibility of models to others. For example, a common pattern amongst Stan users (and in the Stan manual) is to use variables like `J` to refer to the number of elements in a group (e.g., number of participants) and a corresponding index like `j` to refer to specific elements in that group. I believe this sacrifices too much readability for the sake of concision; I prefer a pattern like `n_participant` for the size of the group and `participant` (or a mnemonic shortform like `p`) for specific elements. In functions where names are auto-generated (like `compose_data`), `tidybayes` will (by default) assume you want these sorts of more descriptive names; however, you can always override the default naming scheme.


## Setup

```{r, include=FALSE}
knitr::opts_chunk$set(  #default code chunk options
    dev = "CairoPNG"      #nicer PNG figures
)
```

The following libraries are required to run this vignette:

```{r setup, message=FALSE}
library(magrittr)
library(dplyr)
library(tidybayes)
library(ggplot2)
library(ggstance)
library(rstan)
```

These options help Stan run faster:

```{r, eval=FALSE}
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```

```{r, include=FALSE}
# While the previous code chunk is the actual recommended approach,
# CRAN vignette building policy limits us to 2 cores, so we use at most
# 2 to build this vignette (but show the previous chunk to the reader
# as a best pratice example)
rstan_options(auto_write = TRUE)
options(mc.cores = min(2, parallel::detectCores()))
```

And these options use a simpler ggplot theme:

```{r}
theme_set(theme_light())
```


## Example dataset

```{r eval=FALSE, include=FALSE}
# generate the dataset used below
n = 10
n_condition = 5
ABC =
    data_frame(
        condition = rep(c("A","B","C","D","E"), n),
        y = rnorm(n * 5, c(0,1,2,1,-1), 0.5)
    ) %>%
    arrange(condition)
#save(ABC, file = "data/ABC.Rdata")
```

To demonstrate `tidybayes`, we will use a simple dataset with 10 observations from 5 conditions each:

```{r}
data(ABC, package = "tidybayes")
```

A snapshot of the data looks like this:

```{r}
ABC
```

This is a typical tidy format data frame: one observation per row. Graphically:

```{r}
ABC %>%
    ggplot(aes(x = condition, y = response)) +
    geom_point() 
```

## Using `compose_data` to prepare a data frame for the sampler

Shunting data from a data frame into a format usable in samplers like JAGS or Stan can involve a tedious set of operations, like generating index variables storing the number of operations or the number of levels in a factor. `compose_data` automates these operations.

A hierarchical model of our example data might estimate an overall mean across the conditions (`overall_mean`), the standard deviation of the condition means (`condition_mean_sd`), the mean within each condition (`condition_mean[condition]`) and the standard deviation of the observations given a condition mean (`y_sd`):

```{stan, output.var="ABC_stan"}
data {
    int<lower=1> n;
    int<lower=1> n_condition;
    int<lower=1, upper=n_condition> condition[n];
    real response[n];
}
parameters {
    real overall_mean;
    vector[n_condition] condition_zoffset;
    real<lower=0> y_sd;
    real<lower=0> condition_mean_sd;
}
transformed parameters {
    vector[n_condition] condition_mean;
    condition_mean = overall_mean + condition_zoffset * condition_mean_sd;
}
model {
    y_sd ~ cauchy(0, 1);                # => half-cauchy(0, 1)
    condition_mean_sd ~ cauchy(0, 1);   # => half-cauchy(0, 1)
    overall_mean ~ normal(0, 5);
    condition_zoffset ~ normal(0, 1);   # => condition_mean ~ normal(overall_mean, condition_mean_sd)
    for (i in 1:n) {
        response[i] ~ normal(condition_mean[condition[i]], y_sd);
    }
}
```

We have compiled and loaded this model into the variable `ABC_stan`.

This model expects these variables as input:

* `n`: number of observations
* `n_condition`: number of conditions
* `condition`: a vector of integers indicating the condition of each observation
* `response`: a vector of observations

Our data frame (`ABC`) only has `response` and `condition`, and `condition` is in the wrong format (it is a factor instead of numeric. However, `compose_data` can generate a list containing the above variables in the correct format automatically. It recognizes that `condition` is a factor and converts it to a numeric, adds the `n_condition` variable automatically containing the number of levels in `condition`, and adds the `n` column containing the number of observations (number of rows in the data frame):

```{r}
compose_data(ABC)
```

This makes it easy to skip right to running the model without munging the data yourself:

```{r}
m = sampling(ABC_stan, data = compose_data(ABC), control = list(adapt_delta=0.99))
```

The results look like this:

```{r}
print(m, pars = c("overall_mean", "condition_mean_sd", "condition_mean", "y_sd"))
```


## Gathering samples from a fit in tidy-format using `spread_samples` and `gather_samples`

### Gathering parameter indices into a separate column in a tidy format data frame

Now that we have our results, the fun begins: getting the samples out in a tidy format! The default methods in Stan for extracting samples from the data do so in a nested format:

```{r}
str(extract(m))
```

The `spread_samples` method lets us instead extract samples into a data frame in tidy format, with a `.chain` and `.iteration` column storing the chain and iteration for each row, and the remaining columns corresponding to parameters or parameter indices. The `spread_samples` method accepts any number of column specifications, which can include names for parameters and names for parameter indices. For example, we can extract the `condition_mean` parameter as a tidy data frame, and put the value of its first (and only) index into the `condition` column, using a syntax that directly echoes how we would specify indices of the `condition_mean` parameter in the model itself:

```{r}
m %>%
    spread_samples(condition_mean[condition])
```

### Automatically converting columns and indices back into their original data types

As-is, the resulting parameters don't know anything about where their indices came from. The index of the `condition_mean` parameter was originally derived from the `condition` factor in the `ABC` data frame. But Stan doesn't know this: it is just a numeric index to Stan, so the `condition` column just contains numbers (`1, 2, 3, 4, 5`) instead of the factor levels these numbers correspond to (`"A", "B", "C", "D", "E"`).

We can recover this missing type information by passing the model through `recover_types` before using `spread_samples`. In itself `recover_types` just returns a copy of the model, with some additional attributes that store the type information from the data frame (or other objects) that you pass to it. This doesn't have any useful affect by itself, but `spread_samples` uses this information to convert any column or index back into the data type of the column with the same name in the original data frame. In this example, `spread_samples` recognizes that the `condition` column was a factor with five levels (`"A", "B", "C", "D", "E"`) in the original data frame, and automatically converts it back into a factor:

```{r}
m %>%
    recover_types(ABC) %>%
    spread_samples(condition_mean[condition])
```

Because we often want to make multiple separate calls to `spread_samples`, it is often convenient to decorate the original model using `recover_types` immediately after it has been fit, so we only have to call it once:

```{r}
m %<>% recover_types(ABC)
```

Now we can omit the `recover_types` call before subsequent calls to `spread_samples`.

## Point estimates and intervals

### With simple parameters

`tidybayes` provides a family of functions for generating point estimates and intervals from samples in a tidy format. These functions follow the naming scheme `[mean|median|mode]_[qi|hdi]`, for example, `mean_qi`, `median_qi`, `mode_hdi`, and so on. The first name (before the `_`) indicates the type of point estimate, and the second name indicates the type of interval. `qi` yields a quantile interval (a.k.a. equi-tailed interval or precentile interval) and `hdi` yields a highest (posterior) density interval. Custom estimates or intervals can also be applied using the `point_interval` function.

For example, we might gather the samples corresponding to the overall mean and standard deviation of observations:

```{r}
m %>%
    spread_samples(overall_mean, y_sd)
```

Like with `condition_mean[condition]`, this gives us a tidy data frame. If we want the mean and 95% quantile interval of the parameters, we can apply `mean_qi`:

```{r}
m %>%
    spread_samples(overall_mean, y_sd) %>%
    mean_qi(overall_mean, y_sd)
```

We can specify the columns we want to get means and intervals from, as above, or if we omit the list of columns, `mean_qi` will use every column that is not a grouping column or a special column (one that starts with `.`, like `.chain` or `.iteration`). Thus in the above example, `overall_mean` and `y_sd` are redundant arguments to `mean_qi` because they are also the only columns we gathered from the model. So we can simplify this to:

```{r}
m %>%
    spread_samples(overall_mean, y_sd) %>%
    mean_qi()
```

### With indexed parameters

When we have a parameter with one or more indices, such as `condition_mean`, we can apply `mean_qi` (or other functions in the `point_estimate` family) as we did before:

```{r}
m %>%
    spread_samples(condition_mean[condition]) %>%
    mean_qi()
```

How did `mean_qi` know what to aggregate? Data frames returned by `spread_samples` are automatically grouped by all index variables you pass to it; in this case, that means it groups by `condition`. `mean_qi` respects groups, and calculates the estimates and intervals within all groups. Then, because no columns were passed to `mean_qi`, it acts on the only non-special (`.`-prefixed) and non-group column, `condition_mean`. So the above shortened syntax is equivalent to this more verbose call:

```{r}
m %>%
    spread_samples(condition_mean[condition]) %>%
    group_by(condition) %>%    # this line not necessary (done automatically by spread_samples)
    mean_qi(condition_mean)
```

## Plotting point estimates and intervals

Plotting means and intervals is straightforward using the "pointrange" geom of `ggplot`, by mapping the corresponding columns onto the `y`, `ymin`, and `ymax` aesthetics used by `geom_pointrange`:

```{r}
m %>%
    spread_samples(condition_mean[condition]) %>%
    mean_qi() %>%
    ggplot(aes(x = condition, y = condition_mean, ymin = conf.low, ymax = conf.high)) +
    geom_pointrange()
```


## Interval estimates with multiple probability levels

`mean_qi` and its sister functions can also produce an arbitrary number of probability intervals by setting the `.prob =` argument:

```{r}
m %>%
    spread_samples(condition_mean[condition]) %>%
    mean_qi(.prob = c(.95, .8, .5))
```

The results are in a tidy format: one row per index (`condition`) and probability level (`.prob`). This facilitates plotting. For example, assigning `-.prob` to the `size` aesthetic will show all intervals, making thicker lines correspond to smaller intervals:

```{r}
m %>%
    spread_samples(condition_mean[condition]) %>%
    mean_qi(.prob = c(.95, .8, .5)) %>%
    ggplot(aes(x = condition, y = condition_mean, ymin = conf.low, ymax = conf.high, 
        size = -.prob    # shorter interval => thicker line
    )) +
    geom_pointrange(
        fatten = 2         # smaller point estimate (otherwise it is very large)
    ) +
    scale_size_continuous(
        range = c(1, 2.5),   # default range is c(0, 6) --- makes very thick lines
        guide = FALSE      # no need for a legend on size
    )
```

## Combining variables with different indices in a single tidy format data frame

`spread_samples` supports gathering variables that have different indices. It automatically matches up indices with the same name, and duplicates values as necessary to produce one row per all combination of levels of all indices. For example, we might want to calculate the difference between each condition mean and the overall mean. To do that, we can gather samples from the overall mean and all condition means:

```{r}
m %>% 
    spread_samples(overall_mean, condition_mean[condition])
```

Within each sample, `overall_mean` is repeated as necessary to correspond to every index of `condition_mean`. Thus, the `mutate` function from dplyr can be used to take the differences over all rows, then we can summarize with `mean_qi`:

```{r}
m %>%
    spread_samples(overall_mean, condition_mean[condition]) %>%
    mutate(condition_offset = condition_mean - overall_mean) %>%
    mean_qi(condition_offset)
```

## Making posterior predictions

We can use combinations of variables with difference indices to generate predictions from the model. In this case, we can combine the condition means with the residual standard deviation to generate predictive distributions from the model:

```{r}
m %>%
    spread_samples(condition_mean[condition], y_sd) %>%
    mutate(y_rep = rnorm(n(), condition_mean, y_sd)) %>%
    ggplot(aes(x = y_rep)) +
    stat_density() +
    facet_grid(. ~ condition) + 
    coord_flip()
```

And even summarize these as predictive intervals and compare to the data:

```{r}
m %>%
    spread_samples(condition_mean[condition], y_sd) %>%
    mutate(y_rep = rnorm(n(), condition_mean, y_sd)) %>%
    mean_qi(y_rep, .prob = c(.95, .8, .5)) %>%
    ggplot(aes(x = condition, y = y_rep)) +
    geom_linerange(aes(ymin = conf.low, ymax = conf.high, 
        color = ordered(-.prob)),
        size = 4) +
    geom_point(aes(y = response), data = ABC) +
    scale_color_brewer(guide = FALSE)
```

If this model is well-calibrated, about 95% of the data should be in the outer intervals, 80% in the next-smallest intervals, and 50% in the smallest intervals.

Altogether, data, posterior predictions, and estimates of the means:

```{r}
samples = m %>%
    spread_samples(condition_mean[condition], y_sd)

reps = samples %>%
    mutate(y_rep = rnorm(n(), condition_mean, y_sd)) %>%
    mean_qi(y_rep, .prob = c(.95, .8, .5))

parameters = samples %>%
    mean_qi(condition_mean, .prob = c(.95, .66))
   
ABC %>%
    ggplot(aes(x = condition)) +
    geom_linerange(
        aes(ymin = conf.low, ymax = conf.high, 
            color = ordered(-.prob)),
        size = 4, 
        data = reps) +
    geom_pointrange(
        aes(y = condition_mean, ymin = conf.low, ymax = conf.high, 
            size = -.prob),
        fatten = 1.5, position=position_nudge(x=0.3),
        data = parameters) +
    geom_point(aes(y = response)) +
    scale_color_brewer(guide = FALSE) +
    scale_size_continuous(range = c(1,2), guide = FALSE)
```

## Comparing levels of a factor

If we wish compare the values of `condition_mean` across conditions, `compare_levels` facilitates comparisons of the value of some variable across levels of a factor. By default it computes all pairwise differences:

```{r, fig.width=7}
#N.B. the syntax for compare_levels is experimental and may change
m %>%
    spread_samples(condition_mean[condition]) %>%
    compare_levels(condition_mean, by = condition) %>%
    ggplot(aes(x = condition, y = condition_mean)) +
    geom_violin(scale="width", fill="gray75", color=NA) +
    stat_summary(aes(size=-...prob..), 
        fun.data=median_qi, fun.args=list(.prob=c(.95,.66)), geom="pointrange", fatten=1.25)+
    scale_size_continuous(range=c(0.5,1.5), guide=FALSE) +
    coord_flip()
```

## Alternative estimates and intervals: mean, median, mode; qi, hdi

The `point_interval` family of functions follow the naming scheme `[mean|median|mode]_[qi|hdi]`, and all work in the same way as `mean_qi`: they take a series of names (or expressions calculated on columns) and summarize those columns with the corresponding point estimate (mean, median, or mode) and interval (qi or hdi). `qi` yields a quantile interval (aka equi-tailed interval or precentile interval) and `hdi` yields a highest (posterior) density interval. These can be used in any combination desired. Replacing `mean_qi` with `mode_hdi` in the previous example yields mode and HDI instead of mean and quantile interval:

```{r}
m %>%
    spread_samples(condition_mean[condition]) %>%
    mode_hdi() %>%
    ggplot(aes(x = condition, y = condition_mean, ymin = conf.low, ymax = conf.high)) +
    geom_pointrange()
```

This is probably more noticeable given a skewed distribution, such as might be expected on a scale parameter:

```{r}
sd_samples = m %>% spread_samples(y_sd)
rbind(
    sd_samples %>% mode_hdi(y_sd) %>% mutate(type="mode_hdi"),
    sd_samples %>% mean_qi(y_sd) %>% mutate(type="mean_qi")
) %>%
    ggplot(aes(x = y_sd)) +
    stat_density(aes(y = ..scaled.. * 10), data = sd_samples, fill="gray75") +
    geom_point(aes(y = type)) +
    geom_errorbarh(aes(y = type, xmin = conf.low, xmax = conf.high), height=0)
```

