---
title: "Using Tidy Data with Bayesian Samplers"
author: "Matthew Kay"
date: "`r Sys.Date()`"
output: 
    rmarkdown::html_vignette:
        toc: TRUE
vignette: >
  %\VignetteIndexEntry{Using Tidy Data with Bayesian Samplers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

This vignette introduces the `tidybayes` package, which facilitates the use of tidy data (one observation
per row) with Bayesian samplers in R. 

The default output (and sometimes input) data formats of popular samplers
like JAGS and Stan often don't quite conform to the ideal of "tidy" data. For example, input formats
might expect a list instead of a data frame, and for all variables to be encoded as numeric values
(requiring translation of factors to numeric values and the creation of index variables to store
the number of levels per factor or the number of observations in a data frame). Output formats will
often be in matrix form (requiring conversion for use with libraries like ggplot), and will use
numeric indices (requiring conversion back into factor level names if the you wish to make meaningfully-labelled
plots or estimates). `tidybayes` automates all of these sorts of tasks.

### Philosophy

There are a few core ideas that run through the `tidybayes` API that should (hopefully) make it easy to use:

1. Variable names in models should be descriptive. This means I tend to avoid cryptic (and short) subscripts
in favor of longer (but descriptive) ones. This is a matter of readability and accessibility of models to 
others. For example, a common pattern amongst Stan users (and in the Stan manual) is to variables like `J` to
refer to the number of elements in a group (e.g., number of participants) and a corresponding index like 
`j` to refer to specific elements in that group. I believe this sacrifices too much readability for the sake 
of concision; I prefer a pattern like `n_participant` for the size of the group and `participant` (or a
mnemonic shortform like `p`) for specific elements. 

2. Sensible defaults make life easy. Thus, `tidybayes` will generate group size variables like `n_participant`
automatically given a factor column in a data frame named `participant`. At the same time, this behavior
can be modified as desired (e.g. if you prefer a different naming scheme for indices).

3. Tidy data means no parameter names as values. In contrast to the `ggmcmc` library (which translates sampler
results into a data frame with a `Parameter` and `value` column), The `gather_samples` function in `tidybayes` is designed to result in data frames where the columns are named after parameters of interest and (in some cases) indices of those parameters, as automatically as possible and using a syntax as close to the same way you would refer to those variables in the sampler's language as possible. `tidybayes` does the tedious work of figuring out how to make a data frame look the way you need it to (even if some of your parameters have different indices).

4. `tidybayes` methods should fit into a workflow familiar to users of the `tidyverse` (`dplyr`, `ggplot2`, etc), 
which means fitting into the pipe (`%>%`) workflow, respecting grouped data frames (allowing methods like `mean_qi` to calculate estimates and intervals for many parameters simultaneously), and not reinventing too much of the wheel if it is already made easy by functions provided by those packages (unless it makes for much clearer code for a common idiom).

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(  #default code chunk options
    dev = "CairoPNG"      #nicer PNG figures
)
```

The following libraries are required to run this vignette:

```{r message=FALSE}
library(magrittr)
library(dplyr)
library(forcats)
library(tidybayes)
library(ggplot2)
library(rstan)
```

And thse options help Stan run faster:

```{r}
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```

```{r include=FALSE}
#ggplot options
theme_set(theme_light())
```


## Using `compose_data` to prepare data for the sampler

```{r eval=FALSE, include=FALSE}
# generate the dataset used below
n = 10
n_group = 5
ABC =
    data_frame(
        group = rep(c("A","B","C","D","E"), n),
        y = rnorm(n * 5, c(0,1,2,1,-1), 0.5)
    ) %>%
    arrange(group)
#save(ABC, file = "data/ABC.Rdata")
```

We will load is a simple dataset with 10 observations from 5 groups each:

```{r}
data(ABC, package = "tidybayes")
```

The data looks like this:

```{r}
ABC
```

Which is a typical tidy format: one observation per row. A plot:

```{r}
ABC %>%
    mutate(group = fct_rev(group)) %>% #order groups from top to bottom on y axis (can omit)
    ggplot(aes(x = y, y = group)) +
    # geom_violin(adjust = 3) + 
    geom_point() 
    # coord_flip()
```

A hierarchical model of the above data might estimate an overall mean across the groups (`overall_mean`), the standard deviation of the group means (`group_mean_sd`), the mean within each group (`group_mean[group]`) and the standard deviation of the observations given a group mean (`y_sd`):

```{r, echo = FALSE, results = "asis"}
cat("<div class='sourceCode'><strong style='float:right;padding:1em;'>ABC.stan</strong>
<pre class='sourceCode stan'><code class='sourceCode stan'>")
writeLines(readLines("ABC.stan"))
cat("</code></pre></div>")
```

This model expects input variables `n` (number of observations), `n_group`, (number of groups), `group` (an array of numbers indicating the group of each observation) and `y` (each observation). While our data frame only has `y` and `group` (as a factor instead of a numeric), `compose_data` can generate a list of the above variables automatically. It recognizes that `group` is a factor and converts it to a numeric, and adds the `n_group` variable automatically with the number of levels in `group`, as well as the `n` column containing the number of observations:

```{r}
compose_data(ABC)
```

This makes it easy run the model without munging the data yourself:

```{r}
m = stan("ABC.stan", data = compose_data(ABC), control = list(adapt_delta=0.99))
```

The results look like this:

```{r}
m
```


# Gathering samples from a fit in tidy-format using `gather_samples`

The estimated parameters from Stan don't retain any information about how they were generated. For example, the index of the `group_mean` parameter---which estimates the mean of each group, and which we are likely interested in---was derived from the `group` factor in the `ABC` data frame. But Stan doesn't know this (it is just a numeric index to Stan), so the results give us `group_mean[1]`, `group_mean[2]`, etc.

We can recover this missing information using the `recover_types` function. In itself `recover_types` just returns a copy of the model, decorated so that type information is drawn from the data frame (or other objects) that you pass to it. This won't have any useful effect until we call `gather_samples` later. For now, let's decorate the model with its recovered type information from the original data frame:

```{r}
m %<>% recover_types(ABC)
```

Now we can gather samples from the model in a tidy format. In the simplest example, we might gather samples from parameters with no index, such as the overall mean and the standard deivation of responses:

```{r}
m %>%
    gather_samples(overall_mean, y_sd)
```

This gives us a tidy data frame, with one column per parameter and one sample per row. If we want the mean and 95% quantile interval, we can use mean_qi:

```{r}
m %>%
    gather_samples(overall_mean, y_sd) %>%
    mean_qi(overall_mean, y_sd)
```

What's more interesting, though, is when we have a parameter with one or more indices, such as `group_mean`. We can extract this variable using gather_samples as well, by giving a variable name to become the index column in the resulting data frame

```{r}
m %>%
    gather_samples(group_mean[i])
```

This gives one row per sample per index of group_mean. However, it hasn't recovered the original type information! To do that, we have to specify the same name as the index column in the data frame passed to `recover_types` above. Since the index factor was called `group` in the `ABC` data frame, if we pass that to `gather_samples`, it will automatically recognize this and recover the factor levels in the returned data frame:

```{r}
m %>%
    gather_samples(group_mean[group])
```

And because `mean_qi` (and its sister functions like `mode_hdi`, `median_qi`, etc---see `?point_estimate`) all obey data frame groups, we can get estimates and intervals for each group:

```{r}
m %>%
    gather_samples(group_mean[group]) %>%
    group_by(group) %>%
    mean_qi(group_mean)
```

Then plotting these means is straightforward:

```{r}
m %>%
    gather_samples(group_mean[group]) %>%
    group_by(group) %>%
    mean_qi(group_mean) %>%
    ggplot(aes(x = group, y = group_mean, ymin = group_mean.lower, ymax = group_mean.upper)) +
    geom_pointrange()
```

`mean_qi` and its sister functions can also produce an arbitrary number of probability intervals by setting the `prob =` argument:

```{r}
m %>%
    gather_samples(group_mean[group]) %>%
    group_by(group) %>%
    mean_qi(group_mean, prob = c(.95, .8, .5))
```

Again, in tidy format, facilitating plotting. For example, we can assign the inverse of the `group_mean.prob` column to the `size` aesthetic, making thicker lines corresponding to smaller intervals:

```{r}
m %>%
    gather_samples(group_mean[group]) %>%
    group_by(group) %>%
    mean_qi(group_mean, prob = c(.95, .8, .5)) %>%
    ggplot(aes(x = group, y = group_mean, ymin = group_mean.lower, ymax = group_mean.upper, 
        size = -group_mean.prob    # shorter interval => thicker line
    )) +
    geom_pointrange(
        fatten = 2         # smaller point estimate (otherwise it is very large)
    ) +
    scale_size_continuous(
        range = c(0, 3),   # default range is c(0, 6) --- makes very thick lines
        guide = FALSE      # no need for a legend on size
    )
```



    

```{r}
m %>%
    recover_types(ABC) %>%
    gather_samples(overall_mean, group_mean[group]) %>%
    # mutate(group_mean = overall_mean + group_offset) %>%
    group_by(group) %>%
    mean_qi(group_mean, prob=c(.50,.95)) %>%
    ggplot(aes(x = group, y = group_mean, ymin=group_mean.lower, ymax=group_mean.upper, size=-group_mean.prob)) +
    geom_pointrange(fatten=2, position=position_dodge(width=0.2)) +
    geom_point(aes(y=y, ymin=y, ymax=y), data=ABC, size=1, color="black") +
    scale_size_continuous(range = c(1,2))
```


