---
title: "Dotplot benchmarks"
author: "Matthew Kay"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    df_print: kable
    math_method: mathjax
vignette: >
  %\VignetteIndexEntry{Lineribbon stats and geoms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, child="children/chunk_options.txt"}
```

## Introduction

This article provides benchmarks of some dotplot layout algorithms in ggdist.

## Setup

The following libraries are required to run this vignette:

```{r setup, message = FALSE, warning = FALSE}
library(rlang)
library(distributional)
library(ggdist)
library(ggplot2)
library(ggrepel)
library(dplyr)

theme_set(theme_ggdist())
```

```{r hidden_options, include=FALSE}
.old_options = options(width = 100)
```

## Layouts

We will benchmark using using the primary layout algorithms plus the old implementation of the swarm
algorithm (`layout_oldswarm()`):

```{r layouts}
layouts = list(
  bin = layout_bin(),
  weave = layout_weave(),
  hex = layout_hex(),
  swarm_3 = layout_swarm(strata = 3),
  swarm_inf = layout_swarm(strata = Inf),
  beeswarm = layout_oldswarm()
)
```

## Binwidth finding benchmarks

We'll benchmark finding binwidths on a few different distributions at different sizes:

```{r}
dists = list(
  normal = rnorm,
  uniform = runif,
  exponential = rexp,
  constant = \(n) rep(1, n)
)

set.seed(12345)
df = bench::press(
  layout = names(layouts),
  n = c(100, 1000, 2500, 5000),
  dist = names(dists),
  grouped = c(TRUE, FALSE),
  {
    x = dists[[dist]](n)
    group = if (grouped) rep(1:5, length.out = n) else 1L
    bench::mark(
      find_dotplot_binwidth(x, 1, layout = layouts[[layout]], group = group),
      min_iterations = 20
    )
  }
)
```

Results:

```{r}
df |>
  median_qi(time) |>
  mutate(layout = forcats::fct_reorder(layout, -time)) |>
  ggplot(aes(x = n, y = time)) +
  geom_lineribbon(
    aes(
      color = layout,
      fill = after_scale(alpha(color, 0.25)),
      ymin = .lower,
      ymax = .upper,
      linetype = grouped
    )
  ) +
  geom_point() +
  bench::scale_y_bench_time(base = NULL) +
  coord_cartesian(ylim = c(0, .6)) +
  # scale_y_log10() +
  facet_wrap(~ dist) +
  theme(panel.grid.major.y = element_line(color = "gray75"))
```

## Binning benchmarks

### Small bins

Layout on many small bins can be slow in some pathological cases if we aren't careful in our
implementation (doing splits on bins becomes costly), so we'll benchmark that:

```{r}
x = ppoints(10000)

df = bench::press(
  layout = names(layouts),
  side = c("top", "both"),
  {
    bench::mark(
      bin_dots(x, binwidth = 0.00001, layout = layouts[[layout]], overlaps = "keep", side = side),
      min_iterations = 1
    )
  }
)

df
```

```{r}
df |>
  ggplot(aes(y = layout, x = median, color = side)) +
  geom_point(position = position_dodge(width = 0.2)) +
  guides(color = guide_legend(reverse = TRUE)) +
  bench::scale_x_bench_time(base = NULL)
```
